### 2017-08-17 paper.tex
initial commit

\documentclass{chi-ext}
% Please be sure that you have the dependencies (i.e., additional LaTeX packages) to compile this example.
% See http://personales.upv.es/luileito/chiext/

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
% \copyrightinfo{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
% DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

\title{Gesture Typing on Virtual Tabletop: Effect of Input on Performance}

\numberofauthors{4}
% Notice how author names are alternately typesetted to appear ordered in 2-column format;
% i.e., the first 4 autors on the first column and the other 4 auhors on the second column.
% Actually, it's up to you to strictly adhere to this author notation.
\author{
  \alignauthor{
    \textbf{First Author}\\
    \affaddr{AuthorCo, Inc.}\\
    \affaddr{123 Author Ave.}\\
    \affaddr{Authortown, PA 54321 USA}\\
    \email{author1@anotherco.com}
  }\alignauthor{
    \textbf{Fourth Author}\\
    \affaddr{AuthorCo, Inc.}\\
    \affaddr{123 Author Ave.}\\
    \affaddr{Authortown, PA 54321 USA}\\
    \email{author5@anotherco.com}
  }
  \vfil
  \alignauthor{
    \textbf{Second Author}\\
    \affaddr{AuthorCo, Inc.}\\
    \affaddr{123 Author Ave.}\\
    \affaddr{Authortown, PA 54321 USA}\\
    \email{author2@anotherco.com}
  }\alignauthor{
    \textbf{Fifth Author}\\
    \affaddr{AuthorCo, Inc.}\\
    \affaddr{123 Author Ave.}\\
    \affaddr{Authortown, PA 54321 USA}\\
    \email{author6@anotherco.com}
  }
\vfil
}

% Paper metadata (use plain text, for PDF inclusion and later re-using, if desired)
\def\plaintitle{Gesture Typing on Virtual Tabletop: Effect of Input on Performance}
\def\plainauthor{Name}
\def\plainkeywords{Gesture Input, Gesture Keyboard, Mobile, Continuous Interaction, Tabletop, Text Input}
\def\plaingeneralterms{Documentation, Standardization}

\hypersetup{
  % Your metadata go here
  pdftitle={\plaintitle},
  pdfauthor={\plainauthor},
  pdfkeywords={\plainkeywords},
  pdfsubject={\plaingeneralterms},
  % Quick access to color overriding:
  %citecolor=black,
  %linkcolor=black,
  %menucolor=black,
  %urlcolor=black,
}

\usepackage{graphicx}   % for EPS use the graphics package instead
\usepackage{balance}    % useful for balancing the last columns
\usepackage{bibspacing} % save vertical space in references

\usepackage{subfigure}

% own command
\newcommand{\reffigure}[1]{Figure~\ref{#1}}
\newcommand{\reftable}[1]{Table~\ref{#1}}
\newcommand{\smit}[1]{{\small\textit{{#1}}}}
\newcommand{\cdt}[1]{{\small\uppercase{{#1}}}}
\newcommand{\wpm}{\cdt{wpm} }

\begin{document}

\maketitle

\begin{abstract}
Gesture typing performed through tabletop interaction presents potential for severely/physically impaired users. In this work, we use virtual touch surfaces to bring gesture typing interaction to a tabletop scenario. We first report on the prototype developed and follow with a user study that compares our interaction technique to a control tablet scenario, and explore the influence of input size and aspect ratio on the text input performance. We show that novice user performs at half the rate with our system as compared to control, and that a size of A5/A4 present the best compromise between precision and user preference.
\end{abstract}

\keywords{\plainkeywords}
\textcolor{red}{Optional section to be included in your final version.}

\category{H.5.m}{Information interfaces and presentation (e.g., HCI)}{Miscellaneous}.
%See \cite{ACMCCS}
See: \url{http://www.acm.org/about/class/1998/}
for help using the ACM Classification system.
\textcolor{red}{Optional section to be included in your final version, but strongly encouraged.}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/banner.pdf}
    \caption{Potential interaction setup. A mobile device whose optical sensor creates an on-demand touch surface offers two degrees of freedom in the choice of surface geometry.}~\label{fig:banner}
\end{figure}
\marginpar{
\begin{figure}
  \begin{center}
  \includegraphics[width=\marginparwidth]{figures/pointcloud.pdf}
  \caption{Real segmented sensor image from our prototype with in blue the points within the interactive surface, in cyan the ROI plane, in yellow the detected fingers and in red the user pointer.}
  \label{fig:marginparsample}
  \end{center}
\end{figure}
}
% =============================================================================
\section{Introduction}
% =============================================================================

The commoditization of depth camera creates opportunity for novel interaction scenarios. From its inception with Omnitouch~\cite{Harrison2011} to more recent work, such as Xiao et al. ~\cite{Xiao2016}, using vision-based system for tabletop interaction affords economy and flexible designs. Virtual surfaces are indeed parametrisable, tackle the palm rejection problem and among others allow for dirty hand/sterile scenarios.

This can be interesting for patients with severe physical disabilities. The flexibility of such a system allow for rapid deployment but also let the designer finding the best parameters for the interaction in terms of size and touch point. Moreover, tabletop interaction is nice to the body since it allows to the forearm during the interaction.

Given that text-input remains a very important activity on mobile devices~\cite{McGregor2014}, we set out to explore the feasibility of gesture typing~\cite{Kristensson2004} on the prototype as it suits well our interaction paradigm.

This work aims at making the following two specific contributions:
\begin{itemize}
\item we describe supervised learning solution to the fingertip touch classification problem.
\item we analyse the influence of input space in terms of size and aspect ratio on the gesture typing performance.
\end{itemize}

% =============================================================================
\section{System overview}
% =============================================================================

Before implementing a prototype, we gathered some feedback from a Wizard of Oz testing where 4 participants with high spinal cord injury performed gesture typing like shapes on a tabletop. We gained the insights that this user group is capable of performing the said trajectories, up to A4/A3 and that

The field of hand pose estimation is rapidly moving, and at the time of the study we did not find a software package that would fill all the requirements. We chose to perform simple hand tracking with Kalman filtering of the closest point to the camera, and did spend some time on the touch classification - which (is rarely tackled/has proven problematic) in the litterature~\cite{Xiao2016}.

This problem is usually solved by manually setting threshold from observation from the data. This is however very well suited for a supervised learning scheme. With that in mind, we recorded 12 minutes of film, equally balanced with touching and non-touching frames, as well as different fingers (pinky, middle and index) and different pose angle. We trained a simple MLP classifier.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/roc_auc.pdf}
    \caption{Performance of the fingertip touch classification based on a classic supervised machine learning algorithm.}~\label{fig:banner}
\end{figure}

From the recording camera, and after processing on a stationnary computer (in soft realtime), the detected touch events as sent to a mobile device which runs a custom application and reacts as if the user was touching the screen. An additional audiovisual feedback is presented to display hover position and touch down events.

% =============================================================================
\section{User study}
% =============================================================================
We designed a user study with two goals in mind, evaluating the usiability of the system as compared to an interaction on a mobile device, and finding out the influence of the size and aspect ratio of the input surface onto the text input rate. The design is detailed in table.

We gathered 12 participants, all right handed, only one had previous experience with gesture typing.
We  balanced the LEVEL condition. The task was to write in succession 20 words taken at random from the most common english words with length between 2 and 5 letters\footnote{Word sample used for the experiment: \{\cdt{asset}, \cdt{badly}, \cdt{chew}, \cdt{deck}, \cdt{dog}, \cdt{end}, \cdt{fix}, \cdt{gate}, \cdt{herb}, \cdt{hide}, \cdt{hot}, \cdt{iron}, \cdt{lab}, \cdt{lay}, \cdt{order}, \cdt{safe}, \cdt{seek}, \cdt{seem}, \cdt{want}, \cdt{wrong}\}}. The design of the task allow the user to familiarise with gesture typing and emulates a proficient bahavior. We also asked the user to fill in a NASA Task Load Index.

\begin{table}
  \centering
  \begin{tabular}{l l | r r r r r}
    \smit{LEVEL} & \smit{DEVICE} & \smit{width} & \smit{height} & \smit{area}& \smit{ratio} & \smit{$CD_{gain}$} \\
    % \midrule
    \hline
    \cdt{OP1} & \cdt{optical} & 9.4 & 4.7 & 44.2 & 2 & 1 \\
    \cdt{OP2} & \cdt{optical} & 18.8 & 9.4 & 176.7 & 2 & 1/2 \\
    \cdt{OL2} & \cdt{optical} & 25.6 & 6.9 & 176.6 & 3.7 & 1/1.7 \\
    \cdt{OP4} & \cdt{optical} & 37.7 & 18.9 & 712.5 & 2 & 1/4 \\
    % \midrule
    \hline
    \cdt{tp1} & \cdt{tablet} & 9.4 & 4.7 & 44.2 & 2 & 1 \\
    % \bottomrule
  \end{tabular}
  \caption{Dimensions in centimeters, area in squared centimeter, ratio and control/display gain for all 5 combinations used in the experiment.}~\label{tab:cdt}
\end{table}

% =============================================================================
\section{Preliminary result}
% =============================================================================
We measure the dependent variable time to completion and ...
The dependent variables are the success rate, the time taken per trial and the trace data when available. This allowed us to compute the dependent measure error rate defined as the per- centage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in [3], according to the formula: $WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds.

We found that LAY presented some very unusual behavior. Entry rates had an error rate at $89.5\%$ while the WORD mean is $20.1\%$ and no other word had an error rate higher than $30\%$. The reason is that the recogniser promotes words of higher prior probability in the language model, in this case “Larry”, “last” or “Katy” - highlight the limitations of pure gesture typing interaction.

A statistical analysis showed a significant main effect of DEVICE (F1,11 = 37.77, p < 0.0001) on error rate with mean value for TP1 and OP1 equal to respectively 6.2\% and 26.1\%, see Figure 4.a. A statistical analysis also showed a significant main effect of SIZE (F2,22 = 10.99, p < 0.001) on the error rate. Finally, an ANOVA on TP1, OP2, OP4 and OL2 did not show a significant main effect (p = 0.11) even though OPTICAL is in average higher than TABLET. In other words, among all levels of the experiment, only OP1 shows a significant higher error rate.

A statistical analysis showed a significant main effect of DEVICE (F1,11 = 90.15, p < 0.0001) on input rate with mean values for TP1 and OP1 equal to respectively 29.2 WPM and 13.7 WPM. We also looked at pair-wise comparison2 for OP-TICAL and could not find a statistical difference in the mean input rate. The input rate achieved by the participants in TP1 is in-line with what can be expected from novice users after the time of the experiments~\cite{Kristensson2004}. The averaged 54\% lower input rate in OPTICAL should be compared with other simi- lar results, as~\cite{Markussen2014} with 57\% after 10 sessions, that compare direct and indirect input modality for mid-air gesture typing.

The participants were capable of moving their hand with great variability when it comes to speed.

The Nasa TLX reported high stress on big surfaces, and high stress on the very small surface. As a results, the optimal size is in the middle.

\begin{figure}
\centering     %%% not \center
\subfigure[]{\includegraphics[width=0.49\linewidth]{figures/err_DEVICE_SHAPE.pdf}}
\subfigure[]{\includegraphics[width=0.49\linewidth]{figures/wpm_DEVICE_SHAPE.pdf}}
\caption{Effect of levels on error rate (a) and input rate (b).}~\label{fig:effect_shape}
\end{figure}

% =============================================================================
\section{Conclusion and Future work}
% =============================================================================
The user study showed that the users hit the accuracy ceiling when using a small surface. It also showed that the task behave according to Fitts law.

The user study These encouraging results open the way to testing the system with patients with spinal cord injury. More so, due to recent advance in hand pose recognition, the reality of interacting via touch through vision based system should be investigated further.

\section{Acknowledgements}
We would like to thank the EU Moregrasp project for funding the research, as well as the patients in the spinal unit in Heidelberg for providing early feedback about the interaction technique.

\balance
\bibliographystyle{acm-sigchi}
\bibliography{collection}

\end{document}

### 2017-08-18 paper.tex
tune biblio to take less space, moved all pictures into margins, add gitignore

@@
 % Paper metadata (use plain text, for PDF inclusion and later re-using, if desired)
-\def\plaintitle{Gesture Typing on Virtual Tabletop: Effect of Input on Performance}
+\def\plaintitle{Gesture Typing on Virtual Surfaces: Effect of Input on Performance}
 \def\plainauthor{Name}
@@
 Gesture typing performed through tabletop interaction presents potential for severely/physically impaired users.
-In this work, we use virtual touch surfaces to bring gesture typing interaction to a tabletop scenario.
-We first report on the prototype developed and follow with a user study that compares our interaction technique to a control tablet scenario, and explore the influence of input size and aspect ratio on the text input performance.
-We show that novice user performs at half the rate with our system as compared to control, and that a size of A5/A4 present the best compromise between precision and user preference.
+In this work, we use depth camera to create virtual touch surfaces on regular tabletop.
+We describe our prototype system and report on a supervised learning approach to fingertip touch detection problem.
+We follow with a user study that compares our indirect interaction technique to control tablet scenario and explore the influence of input size and aspect ratio on the text input performance.
+We show that novice user performs with same error rate at half the input rate with our system as compared to control, and that an input size between A5 and A4 present the best tradeoff between precision and user preference.
 \end{abstract}
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{figures/banner.pdf}
-    \caption{Potential interaction setup. A mobile device whose optical sensor creates an on-demand touch surface offers two degrees of freedom in the choice of surface geometry.}~\label{fig:banner}
-\end{figure}
-\marginpar{
-\begin{figure}
-  \begin{center}
-  \includegraphics[width=\marginparwidth]{figures/pointcloud.pdf}
-  \caption{Real segmented sensor image from our prototype with in blue the points within the interactive surface, in cyan the ROI plane, in yellow the detected fingers and in red the user pointer.}
-  \label{fig:marginparsample}
-  \end{center}
-\end{figure}
-}
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\linewidth]{figures/banner.pdf}
+%     \caption{Potential interaction setup. A mobile device whose optical sensor creates an on-demand touch surface offers two degrees of freedom in the choice of surface geometry.}~\label{fig:banner}
+% \end{figure}
+
 % =============================================================================
@@
 
-The commoditization of depth camera creates opportunity for novel interaction scenarios.
-From its inception with Omnitouch~\cite{Harrison2011} to more recent work, such as Xiao et al.
-~\cite{Xiao2016}, using vision-based system for tabletop interaction affords economy and flexible designs.
-Virtual surfaces are indeed parametrisable, tackle the palm rejection problem and among others allow for dirty hand/sterile scenarios.
+Interaction on virtual tabletop is a fast-moving field nowadays with many different enabling technology.
+Depth cameras have been popular~\cite{Harrison2011,Xiao2016}, but other sensors have been used to the same effect (cite Ubitouch and FTIR).
+Some interaction advantages of all or part of mentioned technologies is economic and flexible designs, ability to tackle the palm rejection problem, support for dirty hand/sterile scenarios, relatively low muscle requirement with arm rest.
 
-This can be interesting for patients with severe physical disabilities.
+\marginpar{
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=\marginparwidth]{figures/banner.pdf}
+        \caption{Potential interaction setup. A mobile device whose optical sensor creates an on-demand touch surface offers two degrees of freedom in the choice of surface geometry.}
+        \label{fig:banner}
+    \end{center}
+\end{figure}
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=\marginparwidth]{figures/system_insitu_0.jpg}
+        \caption{Picture of the user study setup. A tablet device provides the audiovisual feedback while a depth camera mounted on a tripod creates a virtual touch input surface.}
+        \label{fig:system_insitu_0}
+    \end{center}
+\end{figure}
+}
+
+Mention that no standard system is available?
+
+More space.
+
+More space.
+
+More space.
+
+More space.
+
+Gesture typing~\cite{Kristensson2004} is interesting to research in this context for several reasons.
+Text-input is still main activity~\cite{McGregor2014} on mobile device, 40\% of the interaction time.
+The technique itself mitigates the importance of touch classification which has shown to be problematic with optical systems, and as an interaction task little if known about the impact of input size (one of the free parameters of virtual surface) on writing performance.
+
+% The commoditization of depth camera creates opportunity for novel interaction scenarios.
+From its inception with Omnitouch to more recent work, such as Xiao et al.
+~\cite{Xiao2016}, using vision-based system for tabletop interaction affords .
+Virtual surfaces are indeed parametrisable,
+
+% This can be interesting for patients with severe physical disabilities.
 The flexibility of such a system allow for rapid deployment but also let the designer finding the best parameters for the interaction in terms of size and touch point.
@@
 
-Given that text-input remains a very important activity on mobile devices~\cite{McGregor2014}, we set out to explore the feasibility of gesture typing~\cite{Kristensson2004} on the prototype as it suits well our interaction paradigm.
+% Given that text-input remains a very important activity on mobile devices~\cite{McGregor2014}, we set out to explore the feasibility of gesture typing on the prototype as it suits well our interaction paradigm.
 
-This work aims at making the following two specific contributions:
+This work specifically focus on two goals:
+% aims at making the following two specific contributions.
+First, show that supervised learning is a good approach to fingertip touch classification problem.
+Second, analyse the influence of input space on gesture typing performance.
 \begin{itemize}
-\item we describe supervised learning solution to the fingertip touch classification problem.
-\item we analyse the influence of input space in terms of size and aspect ratio on the gesture typing performance.
+\item to investigate supervised learning as an approach to fingertip touch classification problem.
+\item to study the influence of input space (size and aspect ration) on gesture typing performance.
 \end{itemize}
@@
 
-Before implementing a prototype, we gathered some feedback from a Wizard of Oz testing where 4 participants with high spinal cord injury performed gesture typing like shapes on a tabletop.
-We gained the insights that this user group is capable of performing the said trajectories, up to A4/A3 and that
+As shown by~\autoref{fig:banner}, the envisioned interaction scenario.
+The implementation is picture in \autoref{fig:system_insitu_0}.
+We implemented a simple finger tracker that will follow the closest point to the camera.
 
-The field of hand pose estimation is rapidly moving, and at the time of the study we did not find a software package that would fill all the requirements.
+In parallel to implementing a prototype, we gathered some informal feedback from a Wizard of Oz testing where 4 participants with high spinal cord injury performed gesture typing like shapes on a tabletop.
+We gained the insights that this user group is capable of performing the said trajectories, up to A4/A3 and even though ...
+(not sure I can write that)
+
+Touch classification is usually solved with flood filling as in.
+We however would mention that this problem is a classical supervised learning problem if some labeled test data can be recorded.
+An author of the current work recorded 12 minutes of video and fed that to a MLP after feature extraction.
+The performance is good with average 96\% AUC.
+This was as usual also fitted with an asymetric touch level.
+
+More space
+
+More space
+
+More space
+
+More space
+% The field of hand pose estimation is rapidly moving, and at the time of the study we did not find a software package that would fill all the requirements.
 We chose to perform simple hand tracking with Kalman filtering of the closest point to the camera, and did spend some time on the touch classification - which (is rarely tackled/has proven problematic) in the litterature~\cite{Xiao2016}.
 
-This problem is usually solved by manually setting threshold from observation from the data.
+% This problem is usually solved by manually setting threshold from observation from the data.
 This is however very well suited for a supervised learning scheme.
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{figures/roc_auc.pdf}
-    \caption{Performance of the fingertip touch classification based on a classic supervised machine learning algorithm.}~\label{fig:banner}
-\end{figure}
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\linewidth]{figures/roc_auc.pdf}
+%     \caption{Performance of the fingertip touch classification based on a classic supervised machine learning algorithm.}~\label{fig:banner}
+% \end{figure}
 
@@
 The design is detailed in table.
+
+\marginpar{
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/pointcloud.pdf}
+    \caption{Real segmented sensor image from our prototype with points within the interactive surface (blue), the ROI plane (cyan), the detected fingers (yellow) and the user pointer (red).}
+    \label{fig:pointcloud}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/roc_auc.pdf}
+    \caption{Performance of the fingertip touch classification based on a classic supervised machine learning algorithm.}~\label{fig:roc_auc}
+\end{figure}
+}
 
@@
 
+More space
+
+More space
+
+More space
+
 % =============================================================================
@@
 
+\marginpar{
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/err_DEVICE_SHAPE.pdf}
+    \caption{Effect of levels on error rate.}
+    \label{fig:err_DEVICE_SHAPE}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/wpm_DEVICE_SHAPE.pdf}
+    \caption{Effect of levels on input rate.}
+    \label{fig:wpm_DEVICE_SHAPE}
+\end{figure}
+}
+
 The participants were capable of moving their hand with great variability when it comes to speed.
@@
 
-\begin{figure}
-\centering     %%% not \center
-\subfigure[]{\includegraphics[width=0.49\linewidth]{figures/err_DEVICE_SHAPE.pdf}}
-\subfigure[]{\includegraphics[width=0.49\linewidth]{figures/wpm_DEVICE_SHAPE.pdf}}
-\caption{Effect of levels on error rate (a) and input rate (b).}~\label{fig:effect_shape}
-\end{figure}
+% \begin{figure}
+% \centering     %%% not \center
+% \subfigure[]{\includegraphics[width=0.49\linewidth]{figures/err_DEVICE_SHAPE.pdf}}
+% \subfigure[]{\includegraphics[width=0.49\linewidth]{figures/wpm_DEVICE_SHAPE.pdf}}
+% \caption{Effect of levels on error rate (a) and input rate (b).}~\label{fig:effect_shape}
+% \end{figure}
 
@@
 % =============================================================================
+
 The user study showed that the users hit the accuracy ceiling when using a small surface. It also showed that the task behave according to Fitts law.
@@
 
+\newpage
+
 \balance

### 2017-08-22 paper.tex
clear up to results

--- 2017-08-17 paper.tex
@@
   \alignauthor{
-    \textbf{First Author}\\
+    \textbf{Antoine Loriette}\\
     \affaddr{AuthorCo, Inc.}\\
@@
     \affaddr{Authortown, PA 54321 USA}\\
-    \email{author1@anotherco.com}
+    \email{antoine.loriette@gmail.com}
   }\alignauthor{
-    \textbf{Fourth Author}\\
+    \textbf{Sebastian Stein}\\
     \affaddr{AuthorCo, Inc.}\\
@@
   \alignauthor{
-    \textbf{Second Author}\\
+    \textbf{Roderick Murray-Smith}\\
     \affaddr{AuthorCo, Inc.}\\
@@
   }\alignauthor{
-    \textbf{Fifth Author}\\
+    \textbf{John H. Williamson}\\
     \affaddr{AuthorCo, Inc.}\\
@@
 \begin{abstract}
-Gesture typing performed through tabletop interaction presents potential for severely/physically impaired users.
-In this work, we use virtual touch surfaces to bring gesture typing interaction to a tabletop scenario.
-We first report on the prototype developed and follow with a user study that compares our interaction technique to a control tablet scenario, and explore the influence of input size and aspect ratio on the text input performance.
-We show that novice user performs at half the rate with our system as compared to control, and that a size of A5/A4 present the best compromise between precision and user preference.
+The association of tabletop interaction with gesture typing presents potential for physically impaired users.
+In this work, we use depth cameras to create touch surfaces on regular tabletops.
+We describe our prototype system and report on a supervised learning approach to fingertips touch classification.
+We follow with a gesture typing study that compares our system with a control tablet scenario and explore the influence of input size and aspect ratio of the virtual surface on the text input performance.
+We show that novice users perform with the same error rate at half the input rate with our system as compared to the control condition, that an input size between A5 and A4 present the best tradeoff between precision, performance and user preference and that user's indirect tracking ability seems to be the overall performance limiting factor.
+
+\textcolor{red}{potential for more space}
+
 \end{abstract}
@@
 \keywords{\plainkeywords}
-\textcolor{red}{Optional section to be included in your final version.}
+% \textcolor{red}{Optional section to be included in your final version.}
 
@@
 %See \cite{ACMCCS}
-See: \url{http://www.acm.org/about/class/1998/}
-for help using the ACM Classification system.
-\textcolor{red}{Optional section to be included in your final version, but strongly encouraged.}
+% See: \url{http://www.acm.org/about/class/1998/}
+% for help using the ACM Classification system.
+% \textcolor{red}{Optional section to be included in your final version, but strongly encouraged.}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{figures/banner.pdf}
-    \caption{Potential interaction setup. A mobile device whose optical sensor creates an on-demand touch surface offers two degrees of freedom in the choice of surface geometry.}~\label{fig:banner}
-\end{figure}
+% =============================================================================
+\section{Introduction}
+% =============================================================================
+The combination of indirect optically tracked input and potentially projected display (“virtual surfaces”) has several interesting properties.
+Virtual surfaces, as opposed to touch screens, are well suited to tackle the palm rejection problem and the occlusion problem, offer a choice of size, aspect ratio and texture of the input space and allow interactions with dirty or wet hands.
+
+Depth cameras~\cite{Harrison2011,Xiao2016} have usually been employed to create such surfaces, even though touch classification remains problematic and few systems are available for all camera types.
+Other work have tried to combine different sensor sources (cite Ubitouch) or other techniques (FTIR?).
+
 \marginpar{
@@
     \begin{center}
-  \includegraphics[width=\marginparwidth]{figures/pointcloud.pdf}
-  \caption{Real segmented sensor image from our prototype with in blue the points within the interactive surface, in cyan the ROI plane, in yellow the detected fingers and in red the user pointer.}
-  \label{fig:marginparsample}
+        \includegraphics[width=\marginparwidth]{figures/banner.pdf}
+        \caption{Potential interaction setup. A mobile device whose optical sensor creates an on-demand touch surface offers width and height as free parameters.}
+        \label{fig:banner}
+    \end{center}
+\end{figure}
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=\marginparwidth]{figures/system_insitu_0.jpg}
+        \caption{Picture of the user study setup. A tablet device provides the audiovisual feedback while a depth camera mounted on a tripod creates a virtual touch input surface.}
+        \label{fig:system_insitu_0}
     \end{center}
@@
 }
-% =============================================================================
-\section{Introduction}
-% =============================================================================
 
-The commoditization of depth camera creates opportunity for novel interaction scenarios.
-From its inception with Omnitouch~\cite{Harrison2011} to more recent work, such as Xiao et al.
-~\cite{Xiao2016}, using vision-based system for tabletop interaction affords economy and flexible designs.
-Virtual surfaces are indeed parametrisable, tackle the palm rejection problem and among others allow for dirty hand/sterile scenarios.
+% \textcolor{red}{potential for more space}
 
-This can be interesting for patients with severe physical disabilities.
-The flexibility of such a system allow for rapid deployment but also let the designer finding the best parameters for the interaction in terms of size and touch point.
-Moreover, tabletop interaction is nice to the body since it allows to the forearm during the interaction.
+% \textcolor{red}{potential for more space}
 
-Given that text-input remains a very important activity on mobile devices~\cite{McGregor2014}, we set out to explore the feasibility of gesture typing~\cite{Kristensson2004} on the prototype as it suits well our interaction paradigm.
+In this context, the task of gesture typing~\cite{Kristensson2004} is interesting to research for the following reasons.
+Text-input remains a major activity~\cite{McGregor2014} on mobile device occupying up to 40\% of the user interaction time, the technique lends itself well to optical systems by limiting the requirement for repetitive target acquisition or touch classification, and little is known about the impact of input size (one of the free parameters of virtual surface) on writing performance.
 
-This work aims at making the following two specific contributions:
-\begin{itemize}
-\item we describe supervised learning solution to the fingertip touch classification problem.
-\item we analyse the influence of input space in terms of size and aspect ratio on the gesture typing performance.
-\end{itemize}
+This work specifically focus on two goals: to report on a supervised learning approach to the fingertip touch classification task and to study the influence of input space (size and aspect ratio) on gesture typing performance.
+% \begin{itemize}
+% \item to report on a supervised learning approach to the fingertip touch classification task.
+% \item to study the influence of input space (size and aspect ratio) on gesture typing performance.
+% \end{itemize}
 
@@
 % =============================================================================
+The envisioned interaction scenario is shown in ~\autoref{fig:banner} with its implementation in \autoref{fig:system_insitu_0}.
+We use an Intel Realsense SR300 obliquely mounted on a tripod that overlooks the interaction surface.
+Finger detection and tracking is performed on a desktop computer.
+The system implements the 3 state button model, with its touch events sent to the Android tablet running a custom application overwriding its input event system.
+In addition, hover state is displayed as a red marker at the pointer’s position, touching is displayed through the Android debugging facility as a cross spanning the screen and continuous touch is displayed as a trace.
+An audio feedback is produced on each touch down event.
 
-Before implementing a prototype, we gathered some feedback from a Wizard of Oz testing where 4 participants with high spinal cord injury performed gesture typing like shapes on a tabletop.
-We gained the insights that this user group is capable of performing the said trajectories, up to A4/A3 and that
+The pointer tracked by the system is defined as the point cloud closest to the camera in the depth direction ($y$ on ~\autoref{fig:pointcloud}); there is no explicit finger or hand modeling performed.
+Instead, the assumption is that the interacting finger is the furthest protruding object from the user.
 
-The field of hand pose estimation is rapidly moving, and at the time of the study we did not find a software package that would fill all the requirements.
-We chose to perform simple hand tracking with Kalman filtering of the closest point to the camera, and did spend some time on the touch classification - which (is rarely tackled/has proven problematic) in the litterature~\cite{Xiao2016}.
+The touch classification problem is usually solved via flood-filling~\cite{Harrison2011,Xiao2016}.
+However, it is well suited to a supervised learning approach provided availability of a training dataset.
+One of the authors video-taped 6 minutes of interactions, half touching and half hovering with some frames out of range while using the index, thumb and pinky at close and long range.
+From the fingertip pointcloud, we use the histogram of z-values as features (providing some orientation and shape invariance) and a neural network classifier.
+We tested the generalisation with different cross validation splits and obtained an averaged 0.96 AUC for the ROC, see~\autoref{fig:roc_auc}.
+The ROC curves led us to choose an operating point of 0.5 for the live system, which we trained on the whole dataset for 75 epochs.
 
-This problem is usually solved by manually setting threshold from observation from the data.
-This is however very well suited for a supervised learning scheme.
-With that in mind, we recorded 12 minutes of film, equally balanced with touching and non-touching frames, as well as different fingers (pinky, middle and index) and different pose angle.
-We trained a simple MLP classifier.
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=\linewidth]{./figures/pointcloud.pdf}
+%   \caption{Insert a caption below each figure.}
+%   \label{fig:sample}
+% \end{figure}
+
+% =============================================================================
+\section{User study}
+% =============================================================================
+We ran an experiment with two research goals in mind.
+First we wanted to evaluate the usability of our system, for this purpose we included a condition with an interaction directly on the tablet.
+Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance (Accot et al.~\cite{Accot2001} showed a U-shaped curve in performance).
+Our independent variable was thus the control surface dimensions, which is effectively changing the control-display gain defined by: $CD_{gain} = V_{display}/V_{control}$.
+
+\marginpar{
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/pointcloud.pdf}
+    \caption{Segmented sensor image with points within the interactive surface (blue), the ROI plane (cyan), the detected fingers (yellow) and the user pointer (red).}
+    \label{fig:pointcloud}
+\end{figure}
 
@@
     \includegraphics[width=\linewidth]{figures/roc_auc.pdf}
-    \caption{Performance of the fingertip touch classification based on a classic supervised machine learning algorithm.}~\label{fig:banner}
+    \caption{Performance of the fingertip touch classification.}~\label{fig:roc_auc}
 \end{figure}
+}
 
-From the recording camera, and after processing on a stationnary computer (in soft realtime), the detected touch events as sent to a mobile device which runs a custom application and reacts as if the user was touching the screen.
-An additional audiovisual feedback is presented to display hover position and touch down events.
-
-% =============================================================================
-\section{User study}
-% =============================================================================
-We designed a user study with two goals in mind, evaluating the usiability of the system as compared to an interaction on a mobile device, and finding out the influence of the size and aspect ratio of the input surface onto the text input rate.
-The design is detailed in table.
-
-We gathered 12 participants, all right handed, only one had previous experience with gesture typing.
-We  balanced the LEVEL condition.
-The task was to write in succession 20 words taken at random from the most common english words with length between 2 and 5 letters\footnote{Word sample used for the experiment: \{\cdt{asset}, \cdt{badly}, \cdt{chew}, \cdt{deck}, \cdt{dog}, \cdt{end}, \cdt{fix}, \cdt{gate}, \cdt{herb}, \cdt{hide}, \cdt{hot}, \cdt{iron}, \cdt{lab}, \cdt{lay}, \cdt{order}, \cdt{safe}, \cdt{seek}, \cdt{seem}, \cdt{want}, \cdt{wrong}\}}.
-The design of the task allow the user to familiarise with gesture typing and emulates a proficient bahavior.
-We also asked the user to fill in a NASA Task Load Index.
+We gathered 12 participants, all right handed, without requirement on gesture typing experience.
+A repeated measures within-subjects design was used.
+There were 3 conditions and 5 level combinations were evaluated, see \autoref{tab:cdt} for details.
+The presentation order of the 5 combinations was randomised and presented a uniform distribution across the participants.
+The task was to write 20 words per level (words taken at random from the most common english words with length between 2 and 5 letters) with maximum 7 attempt to complete 5 correct input for each word.
+The design of the task is similar to \cite{Quinn2016} and allows novice users to focus on the physical execution instead of the shape recollection, effectively emulating a proficient bahavior even for novice users.
+After each level, participants were offered to take a break before moving to the next one.
+Finally, participants were asked for their feedback using the NASA Task Load Index~\cite{Hart1988} (NASA-TLX) to assess the perceived workload of completed level.
 
@@
     \cdt{tp1} & \cdt{tablet} & 9.4 & 4.7 & 44.2 & 2 & 1 \\
-    % \bottomrule
   \end{tabular}
-  \caption{Dimensions in centimeters, area in squared centimeter, ratio and control/display gain for all 5 combinations used in the experiment.}~\label{tab:cdt}
+  \caption{Design of the experiment. Level name, device type, dimensions (in $cm$), area (in $cm^2$), ratio and control/display gain for all 5 combinations used in the experiment.}~\label{tab:cdt}
 \end{table}
 
+The experimental design was thus: 12 participants $\times$ 5 \cdt{level} $\times$ 20 \cdt{word} = 1200 trials.
+For each trial, we had 5 to 7 attemps depending on the error rate, which means a total of 6,000 to 8,400 total samples.
+After the experiment, we actually recorded 6963 samples.
+
 % =============================================================================
-\section{Preliminary result}
+\section{Result}
 % =============================================================================
-We measure the dependent variable time to completion and ...
 The dependent variables are the success rate, the time taken per trial and the trace data when available.
-This allowed us to compute the dependent measure error rate defined as the per- centage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in [3], according to the formula: $WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds.
+This allowed us to compute the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in [3], according to the formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.
 
-We found that LAY presented some very unusual behavior.
+We found one input word (\cdt{lay}) presented some very unusual behavior.
 Entry rates had an error rate at $89.5\%$ while the WORD mean is $20.1\%$ and no other word had an error rate higher than $30\%$.
@@
 
-A statistical analysis showed a significant main effect of DEVICE (F1,11 = 37.77, p < 0.0001) on error rate with mean value for TP1 and OP1 equal to respectively 6.2\% and 26.1\%, see Figure 4.a.
-A statistical analysis also showed a significant main effect of SIZE (F2,22 = 10.99, p < 0.001) on the error rate.
+A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77, p < 0.0001$) on error rate with mean value for TP1 and OP1 equal to respectively 6.2\% and 26.1\%, see Figure 4.a.
+A statistical analysis also showed a significant main effect of SIZE ($F_{2,22} = 10.99, p < 0.001$) on the error rate.
 Finally, an ANOVA on TP1, OP2, OP4 and OL2 did not show a significant main effect (p = 0.11) even though OPTICAL is in average higher than TABLET.
@@
 
-A statistical analysis showed a significant main effect of DEVICE (F1,11 = 90.15, p < 0.0001) on input rate with mean values for TP1 and OP1 equal to respectively 29.2 WPM and 13.7 WPM.
+A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 90.15, p < 0.0001$) on input rate with mean values for TP1 and OP1 equal to respectively 29.2 WPM and 13.7 WPM.
 We also looked at pair-wise comparison2 for OP-TICAL and could not find a statistical difference in the mean input rate.
@@
 % =============================================================================
-The user study showed that the users hit the accuracy ceiling when using a small surface. It also showed that the task behave according to Fitts law.
 
-The user study These encouraging results open the way to testing the system with patients with spinal cord injury.
+\marginpar{
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/err_DEVICE_SHAPE.pdf}
+    \caption{Effect of levels on error rate.}
+    \label{fig:err_DEVICE_SHAPE}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{figures/wpm_DEVICE_SHAPE.pdf}
+    \caption{Effect of levels on input rate.}
+    \label{fig:wpm_DEVICE_SHAPE}
+\end{figure}
+}
+The user study showed that the users hit the accuracy ceiling when using a small surface.
+It also showed that the task behave according to Fitts law.
+The users seem to be limited by their ability to drive an indirect pointer rather than performing the difficulty of the task itself.
+
+These encouraging results open the way to testing the system with patients with spinal cord injury.
 More so, due to recent advance in hand pose recognition, the reality of interacting via touch through vision based system should be investigated further.
 
-\section{Acknowledgements}
-We would like to thank the EU Moregrasp project for funding the research, as well as the patients in the spinal unit in Heidelberg for providing early feedback about the interaction technique.
+% \section{Acknowledgements}
+% We would like to thank the EU Moregrasp project for funding the research, as well as the patients in the spinal unit in Heidelberg for providing early feedback about the interaction technique.
+
+\newpage
 

### 2017-08-22 paper.tex
up to discussion, missing conclusion, trim citations format

@@
     \textbf{Antoine Loriette}\\
-    \affaddr{AuthorCo, Inc.}\\
-    \affaddr{123 Author Ave.}\\
-    \affaddr{Authortown, PA 54321 USA}\\
+    \affaddr{Dept. Computing Science}\\
+    \affaddr{University of Glasgow}\\
+    \affaddr{Scotland}\\
     \email{antoine.loriette@gmail.com}
@@
     \textbf{Sebastian Stein}\\
-    \affaddr{AuthorCo, Inc.}\\
-    \affaddr{123 Author Ave.}\\
-    \affaddr{Authortown, PA 54321 USA}\\
+    \affaddr{Dept. Computing Science}\\
+    \affaddr{University of Glasgow}\\
+    \affaddr{Scotland}\\
     \email{author5@anotherco.com}
@@
     \textbf{Roderick Murray-Smith}\\
-    \affaddr{AuthorCo, Inc.}\\
-    \affaddr{123 Author Ave.}\\
-    \affaddr{Authortown, PA 54321 USA}\\
+    \affaddr{Dept. Computing Science}\\
+    \affaddr{University of Glasgow}\\
+    \affaddr{Scotland}\\
     \email{author2@anotherco.com}
   }\alignauthor{
-    \textbf{John H. Williamson}\\
-    \affaddr{AuthorCo, Inc.}\\
-    \affaddr{123 Author Ave.}\\
-    \affaddr{Authortown, PA 54321 USA}\\
-    \email{author6@anotherco.com}
+    \textbf{John Williamson}\\
+    \affaddr{Dept. Computing Science}\\
+    \affaddr{University of Glasgow}\\
+    \affaddr{Scotland}\\
+    \email{jhw@dcs.gla.ac.uk}
   }
@@
 We follow with a gesture typing study that compares our system with a control tablet scenario and explore the influence of input size and aspect ratio of the virtual surface on the text input performance.
-We show that novice users perform with the same error rate at half the input rate with our system as compared to the control condition, that an input size between A5 and A4 present the best tradeoff between precision, performance and user preference and that user's indirect tracking ability seems to be the overall performance limiting factor.
+We show that novice users perform with the same error rate at half the input rate with our system as compared to the control condition, that an input size between A5 and A4 present the best tradeoff between performance and user preference and that user's indirect tracking ability seems to be the overall performance limiting factor.
 
@@
 
-Depth cameras~\cite{Harrison2011,Xiao2016} have usually been employed to create such surfaces, even though touch classification remains problematic and few systems are available for all camera types.
-Other work have tried to combine different sensor sources (cite Ubitouch) or other techniques (FTIR?).
+Depth cameras~\cite{Harrison2011,Xiao2016,Sridhar2017} have usually been employed to create such surfaces, while other work have tried to combine different sensor sources~\cite{Wen2016}.
+However, the touch classification, critical to the quality of the interaction remains problematic (sub capacitive quality?) and few systems are available for all camera types.
 
@@
 The system implements the 3 state button model, with its touch events sent to the Android tablet running a custom application overwriding its input event system.
-In addition, hover state is displayed as a red marker at the pointer’s position, touching is displayed through the Android debugging facility as a cross spanning the screen and continuous touch is displayed as a trace.
+In addition, the tablet displays hover state as a red marker at the pointer’s position, touching is displayed through the Android debugging facility as a cross spanning the screen and continuous touch is displayed as a segment line.
 An audio feedback is produced on each touch down event.
@@
 
-The touch classification problem is usually solved via flood-filling~\cite{Harrison2011,Xiao2016}.
+The touch classification problem is usually solved via flood-filling~\cite{Harrison2011,Xiao2016,Sridhar2017}.
 However, it is well suited to a supervised learning approach provided availability of a training dataset.
@@
 First we wanted to evaluate the usability of our system, for this purpose we included a condition with an interaction directly on the tablet.
-Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance (Accot et al.~\cite{Accot2001} showed a U-shaped curve in performance).
-Our independent variable was thus the control surface dimensions, which is effectively changing the control-display gain defined by: $CD_{gain} = V_{display}/V_{control}$.
+Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance (Accot et al.~\cite{Accot2001} demonstrated a U-shaped curve in performance against input scale).
+Our independent variable was thus the control surface dimensions, which is effectively changing the control-display gain ($CD_{gain}$) defined as the ratio between $V_{display}$ and $V_{control}$.
 
@@
 A repeated measures within-subjects design was used.
-There were 3 conditions and 5 level combinations were evaluated, see \autoref{tab:cdt} for details.
+There were 3 conditions and 5 level combinations were evaluated, see \autoref{tab:cdt} below for details.
 The presentation order of the 5 combinations was randomised and presented a uniform distribution across the participants.
+
 The task was to write 20 words per level (words taken at random from the most common english words with length between 2 and 5 letters) with maximum 7 attempt to complete 5 correct input for each word.
@@
 Finally, participants were asked for their feedback using the NASA Task Load Index~\cite{Hart1988} (NASA-TLX) to assess the perceived workload of completed level.
+
+The experimental design was thus: 12 participants $\times$ 5 \cdt{level} $\times$ 20 \cdt{word} = 1200 trials.
+For each trial, we had 5 to 7 attemps depending on the error rate, which means a total of 6,000 to 8,400 total samples.
+After the experiment, we actually recorded 6963 samples.
 
@@
 
-The experimental design was thus: 12 participants $\times$ 5 \cdt{level} $\times$ 20 \cdt{word} = 1200 trials.
-For each trial, we had 5 to 7 attemps depending on the error rate, which means a total of 6,000 to 8,400 total samples.
-After the experiment, we actually recorded 6963 samples.
-
 % =============================================================================
@@
 The dependent variables are the success rate, the time taken per trial and the trace data when available.
-This allowed us to compute the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in [3], according to the formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.
+This allowed us to compute the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in [3], according to the following formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.
 
 We found one input word (\cdt{lay}) presented some very unusual behavior.
-Entry rates had an error rate at $89.5\%$ while the WORD mean is $20.1\%$ and no other word had an error rate higher than $30\%$.
-The reason is that the recogniser promotes words of higher prior probability in the language model, in this case “Larry”, “last” or “Katy” - highlight the limitations of pure gesture typing interaction.
+Its error rate was at $89.5\%$ while the \cdt{WORD} mean is $20.1\%$ and no other word had an error rate higher than $30\%$.
+The reason is that the recogniser promotes words of higher prior probability in the language model, \textit{``Larry''}, \textit{``last''} or \textit{``Katy''}.
+For the all subsequent analysis of error rate, \cdt{lay} is removed from the dataset.
 
-A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77, p < 0.0001$) on error rate with mean value for TP1 and OP1 equal to respectively 6.2\% and 26.1\%, see Figure 4.a.
-A statistical analysis also showed a significant main effect of SIZE ($F_{2,22} = 10.99, p < 0.001$) on the error rate.
-Finally, an ANOVA on TP1, OP2, OP4 and OL2 did not show a significant main effect (p = 0.11) even though OPTICAL is in average higher than TABLET.
-In other words, among all levels of the experiment, only OP1 shows a significant higher error rate.
-
-A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 90.15, p < 0.0001$) on input rate with mean values for TP1 and OP1 equal to respectively 29.2 WPM and 13.7 WPM.
-We also looked at pair-wise comparison2 for OP-TICAL and could not find a statistical difference in the mean input rate.
-The input rate achieved by the participants in TP1 is in-line with what can be expected from novice users after the time of the experiments~\cite{Kristensson2004}.
-The averaged 54\% lower input rate in OPTICAL should be compared with other simi- lar results, as~\cite{Markussen2014} with 57\% after 10 sessions, that compare direct and indirect input modality for mid-air gesture typing.
-
-The participants were capable of moving their hand with great variability when it comes to speed.
-
-The Nasa TLX reported high stress on big surfaces, and high stress on the very small surface. As a results, the optimal size is in the middle.
-
-% \begin{figure}
-% \centering     %%% not \center
-% \subfigure[]{\includegraphics[width=0.49\linewidth]{figures/err_DEVICE_SHAPE.pdf}}
-% \subfigure[]{\includegraphics[width=0.49\linewidth]{figures/wpm_DEVICE_SHAPE.pdf}}
-% \caption{Effect of levels on error rate (a) and input rate (b).}~\label{fig:effect_shape}
-% \end{figure}
-
-% =============================================================================
-\section{Conclusion and Future work}
-% =============================================================================
+A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77, p < 0.0001$) on error rate with mean value for \cdt{TP1} and \cdt{OP1} equal to respectively 6.2\% and 26.1\%, see \autoref{fig:err_DEVICE_SHAPE}.
+A statistical analysis also showed a significant main effect of \cdt{SIZE} ($F_{2,22} = 10.99, p < 0.001$) on the error rate.
+Finally, an ANOVA on \cdt{TP1}, \cdt{OP2}, \cdt{OP4} and \cdt{OL2} did not show a significant main effect ($p = 0.11$) even though \cdt{OPTICAL} is on average higher than \cdt{TABLET}.
+In other words, among all levels of the experiment, only \cdt{OP1} shows a significant higher error rate.
 
@@
 }
-The user study showed that the users hit the accuracy ceiling when using a small surface.
+
+A statistical analysis showed a significant main effect of \cdt{DEVICE} ($F_{1,11} = 90.15, p < 0.0001$) on input rate, see \autoref{fig:wpm_DEVICE_SHAPE}, with mean values for \cdt{TP1} and \cdt{OP1} equal to respectively $29.2 WPM$ and $13.7 WPM$.
+We also looked at pairwise comparison\footnote{Post-hoc analysis were adjusted using Bonferroni correction.} for \cdt{OPTICAL} and could not find a statistical difference in the mean input rate.
+The input rate achieved by the participants in \cdt{TP1} is in-line with what can be expected from novice users after the time of the experiments~\cite{Kristensson2004}.
+The averaged 54\% lower input rate in \cdt{OPTICAL} should be compared with other similar results, as~\cite{Markussen2014} with 57\% after 10 sessions, that compare direct and indirect input modality for mid-air gesture typing.
+
+\subsection{Effect of orientation}
+The design of the experiment also includes \cdt{ORIENTATION}.
+This condition has so far been excluded from the SIZE analysis and can be difficult to apprehend since not only the \cdt{ORIENTATION} of the visual feedback is different but its size also.
+The main result is the average pointer speed in display space at $303.75 pixel/s$, higher than the PORTRAIT orientation at an average $228 pixel/s$.
+It is important to keep in mind that the keyboard area in LANDSCAPE is $61cm^2$, bigger than PORTRAIT at $44cm^2$, which can explain the faster pointer speed but is not responsible for a higher input rate.
+
+\subsection{Qualitative data}
+At the end of the experiment, participants were invited to provide some informal feedback, rank the different level in order of preference and fill in the NASA-TLX form, the tablet interaction was ranked best by all the participants except one.
+\cdt{OP4} was consistently ranked last, while \cdt{OP2} and \cdt{OL2} had equal ranking in second position.
+The NASA-TLX data shows that participants describe \cdt{OP4} as the most physically demanding interaction.
+Participants graded \cdt{OP1} and \cdt{OP4} 20\% higher than \cdt{OP2} and \cdt{OL2} on the scale of effort and frustration.
+Finally, \cdt{OP1} was graded as having lowest level of performance and highest mental and temporal demand.
+
+\section{Discussion}
+
+We proposed a machine learning approach to fingertip touch detection.
+This problem is traditionally addressed by hand-tuning parameters and thresholding.
+In contrast, we showed that using a machine learning approach can solve the problem and generalise well enough to be used in an experiment with 12 unseen participants.
+One improvement could be to delegate the feature extraction step (an histogram of z-values in our case) to a model capable of inferring features directly, such as a convolutional neural network.
+Since the data is cheap to record, we could see this approach being used for many specific cases.
+
+We conducted a user study investigating the usability of the system and showed that for control dimension at least twice the display dimension novice users were capable the same error rate at half the input rate.
+We also showed that at the same $CD_{gain}$, the error rates were dramatically higher.
+Also, the experiment showed a constant input rate across display sizes and aspect ratios.
+Accot et al.~\cite{Accot2001} showed that task with high index of difficulty would exhibits a u-shaped curved with size.
+Since we did not observe such a behavior, this puts in question whether gesture typing for novice users is “hard enough”.
+
+The constant input rate however shows that participants are capable of adapting their motor speed across all investigated scales.
+Participants also varied the display pointer speed, especially in \cdt{OL2} when presented with a bigger display surface.
+Because participants are capable of adapting their motor behavior and their control bahavior, another explanation for the upper bound in text input is that the indirect nature of the interaction is the limiting factor.
+(more research in that direction?)
+
+From the participants feedback, it shows that \cdt{OP1} was a demanding interaction.
+It also showed that increasing the surface size to the extent of \cdt{OP4} is physically demanding.
+Recommended sizes for interaction will thus be \cdt{OP2} or \cdt{OL2} (note that the performance is also dependent on the $CD_{gain}$).
+Moreover, some participants reported the texture having an impact of the interaction.
+Levesque et al.~\cite{Levesque2011} have used friction in a dynamic manner to improve target acquisition.
+Given the relatively low performance of the participants, it could be interesting to explore if tactile feedback could help controlling an indirect pointer.
+
+% % =============================================================================
+% \section{Conclusion and Future work}
+% % =============================================================================
+
+% The user study showed that the users hit the accuracy ceiling when using a small surface.
 It also showed that the task behave according to Fitts law.
@@
 
-These encouraging results open the way to testing the system with patients with spinal cord injury.
+% These encouraging results open the way to testing the system with patients with spinal cord injury.
 More so, due to recent advance in hand pose recognition, the reality of interacting via touch through vision based system should be investigated further.
@@
 
-\newpage
+% \newpage
 
-\balance
+% \balance
 \bibliographystyle{acm-sigchi}

### 2017-08-22 paper.tex
add conclusion, and poster in poster directory

@@
     \affaddr{Scotland}\\
-    \email{antoine.loriette@gmail.com}
+    \email{antoine.loriette@glasgow.ac.uk}
   }\alignauthor{
@@
     \affaddr{Scotland}\\
-    \email{author5@anotherco.com}
+    \email{sebastian.stein@glasgow.ac.uk}
   }
@@
     \affaddr{Scotland}\\
-    \email{author2@anotherco.com}
+    \email{rod@dcs.gla.ac.uk}
   }\alignauthor{
@@
 The dependent variables are the success rate, the time taken per trial and the trace data when available.
-This allowed us to compute the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in [3], according to the following formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.
+This allowed us to compute the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in~\cite{Markussen2014}, according to the following formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.
 
@@
 
-% % =============================================================================
-% \section{Conclusion and Future work}
-% % =============================================================================
-
-% The user study showed that the users hit the accuracy ceiling when using a small surface.
-It also showed that the task behave according to Fitts law.
-The users seem to be limited by their ability to drive an indirect pointer rather than performing the difficulty of the task itself.
-
-% These encouraging results open the way to testing the system with patients with spinal cord injury.
-More so, due to recent advance in hand pose recognition, the reality of interacting via touch through vision based system should be investigated further.
+% =============================================================================
+\section{Conclusion}
+% =============================================================================
+We developed a system that affords gesture typing on arbitrary flat surfaces using depth camera tracking, and demonstrated a machine learning approach to the issue of detecting touch for fingertips which is effective for this mode of control.
+We designed an experiment with two research goals in mind.
+First, we compared the typing performance of our system against a control condition which was a direct interaction on a tablet.
+Second, we studied the influence of size and aspect ratio of the input surface on gesture typing performance.
+We showed that the participants could enter text at half the input rate and the same error rate on surface at least twice the size of the current visual feedback.
+We also showed that the input rate is largely independent of the surface size across the range of sizes we were able to examine.
+Gesture typing has promise for interaction outside of mobile devices (e.g.
+for motor impaired users who struggle with capacitive touch technologies).
+This paper indicates that while camera-tracked gesture typing performance is usable, input rates are lower than touchscreen performance, and this is not influenced by scaling of the input.
 
@@
 
-% \balance
+\balance
 \bibliographystyle{acm-sigchi}

### 2017-08-22 paper.tex
add dimension to title

@@
 
-\title{Gesture Typing on Virtual Tabletop: Effect of Input on Performance}
+\title{Gesture Typing on Virtual Tabletop: Effect of Input Dimension on Performance}
 
@@
 % Paper metadata (use plain text, for PDF inclusion and later re-using, if desired)
-\def\plaintitle{Gesture Typing on Virtual Surfaces: Effect of Input on Performance}
+\def\plaintitle{Gesture Typing on Virtual Surfaces: Effect of Input Dimension on Performance}
 \def\plainauthor{Name}

### 2017-08-22 paper.tex
Update on Overleaf.

@@
     \affaddr{University of Glasgow}\\
-    \affaddr{Scotland}\\
+    \affaddr{Scotland, UK}\\
     \email{antoine.loriette@glasgow.ac.uk}
@@
     \affaddr{University of Glasgow}\\
-    \affaddr{Scotland}\\
+    \affaddr{Scotland, UK}\\
     \email{sebastian.stein@glasgow.ac.uk}
@@
     \affaddr{University of Glasgow}\\
-    \affaddr{Scotland}\\
+    \affaddr{Scotland, UK}\\
     \email{rod@dcs.gla.ac.uk}
@@
     \affaddr{University of Glasgow}\\
-    \affaddr{Scotland}\\
+    \affaddr{Scotland, UK}\\
     \email{jhw@dcs.gla.ac.uk}
@@
 \def\plainauthor{Name}
-\def\plainkeywords{Gesture Input, Gesture Keyboard, Mobile, Continuous Interaction, Tabletop, Text Input}
+\def\plainkeywords{Gesture Input, Gesture Keyboard, Mobile, Continuous Interaction, Tabletop, Text Input, Indirect Interaction}
 \def\plaingeneralterms{Documentation, Standardization}
@@
 \begin{abstract}
-The association of tabletop interaction with gesture typing presents potential for physically impaired users.
+The association of tabletop interaction with gesture typing presents potential for physically and situationally impaired users.
 In this work, we use depth cameras to create touch surfaces on regular tabletops.
-We describe our prototype system and report on a supervised learning approach to fingertips touch classification.
+We describe our prototype system and report on a supervised learning approach to fingertip touch classification.
 We follow with a gesture typing study that compares our system with a control tablet scenario and explore the influence of input size and aspect ratio of the virtual surface on the text input performance.
-We show that novice users perform with the same error rate at half the input rate with our system as compared to the control condition, that an input size between A5 and A4 present the best tradeoff between performance and user preference and that user's indirect tracking ability seems to be the overall performance limiting factor.
+We show that novice users performed with the same error rate at half the input rate with our system as compared to the control condition.
+An input size between A5 and A4 presented the best tradeoff between performance and user preference, and users' indirect tracking ability appears to be the overall performance limiting factor.
 
@@
 The combination of indirect optically tracked input and potentially projected display (“virtual surfaces”) has several interesting properties.
-Virtual surfaces, as opposed to touch screens, are well suited to tackle the palm rejection problem and the occlusion problem, offer a choice of size, aspect ratio and texture of the input space and allow interactions with dirty or wet hands.
+Virtual surfaces, as opposed to touch screens, are well suited to tackle the palm rejection problem and the occlusion problem, offer a choice of size, aspect ratio and texture of the input space, and allow interactions with dirty or wet hands.
 
-Depth cameras~\cite{Harrison2011,Xiao2016,Sridhar2017} have usually been employed to create such surfaces, while other work have tried to combine different sensor sources~\cite{Wen2016}.
-However, the touch classification, critical to the quality of the interaction remains problematic (sub capacitive quality?) and few systems are available for all camera types.
+Depth cameras~\cite{Harrison2011,Xiao2016,Sridhar2017} have been commonly employed to create virtual surfaces, while other works have combined multiple different sensor sources~\cite{Wen2016}.
+Touch classification is critical to the quality of the interaction but remains problematic (sub capacitive quality?) and few systems are available for all camera types.
 
@@
 
-In this context, the task of gesture typing~\cite{Kristensson2004} is interesting to research for the following reasons.
-Text-input remains a major activity~\cite{McGregor2014} on mobile device occupying up to 40\% of the user interaction time, the technique lends itself well to optical systems by limiting the requirement for repetitive target acquisition or touch classification, and little is known about the impact of input size (one of the free parameters of virtual surface) on writing performance.
+In this context, the task of gesture typing~\cite{Kristensson2004} is an interesting research focus.
+Text-input remains a major activity~\cite{McGregor2014} on mobile devices occupying up to 40\% of the user interaction time.
+The technique lends itself well to optical systems by limiting the requirement for repetitive target acquisition or touch classification, yet little is known about the impact of input size (one of the free parameters of virtual surface) on writing performance.
 
-This work specifically focus on two goals: to report on a supervised learning approach to the fingertip touch classification task and to study the influence of input space (size and aspect ratio) on gesture typing performance.
+This work specifically focuses on two goals: to report on a supervised learning approach to the fingertip touch classification task and to study the influence of input space (size and aspect ratio) on gesture typing performance using virtual tabletops.
 % \begin{itemize}
@@
 % =============================================================================
-\section{System overview}
+\section{System Overview}
 % =============================================================================
@@
 Finger detection and tracking is performed on a desktop computer.
-The system implements the 3 state button model, with its touch events sent to the Android tablet running a custom application overwriding its input event system.
+The system implements the 3 state button model, with its touch events sent to the Android tablet running a custom application overwriting its input event system.
 In addition, the tablet displays hover state as a red marker at the pointer’s position, touching is displayed through the Android debugging facility as a cross spanning the screen and continuous touch is displayed as a segment line.
@@
 
+% =============================================================================
+\section{Touch Classification}
+% =============================================================================
 The touch classification problem is usually solved via flood-filling~\cite{Harrison2011,Xiao2016,Sridhar2017}.
@@
 % =============================================================================
-\section{User study}
+\section{User Study}
 % =============================================================================
 We ran an experiment with two research goals in mind.
-First we wanted to evaluate the usability of our system, for this purpose we included a condition with an interaction directly on the tablet.
-Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance (Accot et al.~\cite{Accot2001} demonstrated a U-shaped curve in performance against input scale).
-Our independent variable was thus the control surface dimensions, which is effectively changing the control-display gain ($CD_{gain}$) defined as the ratio between $V_{display}$ and $V_{control}$.
+First we wanted to evaluate the usability of our system.
+For this purpose, we included a condition with an interaction directly on the tablet.
+Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance.
+Accot et al.~\cite{Accot2001} demonstrated a U-shaped curve in performance against input scale.
+Our independent variables were the control surface dimensions, which effectively change the control-display gain ($CD_{gain}$) defined as the ratio between $V_{display}$ and $V_{control}$.
 
@@
 
-We gathered 12 participants, all right handed, without requirement on gesture typing experience.
+We recruited 12 participants, all right handed, without requirement on gesture typing experience.
 A repeated measures within-subjects design was used.
@@
 
-The task was to write 20 words per level (words taken at random from the most common english words with length between 2 and 5 letters) with maximum 7 attempt to complete 5 correct input for each word.
-The design of the task is similar to \cite{Quinn2016} and allows novice users to focus on the physical execution instead of the shape recollection, effectively emulating a proficient bahavior even for novice users.
+The task in each level was to write 20 words; five correct inputs per word with up to seven attempts.
+Words were sampled randomly from the most common English words with two to five letters.
+This task design is inspired by \cite{Quinn2016} and allows novice users to focus on physical execution instead of shape recollection, emulating a proficient bahavior for novice users.
 After each level, participants were offered to take a break before moving to the next one.
@@
 The experimental design was thus: 12 participants $\times$ 5 \cdt{level} $\times$ 20 \cdt{word} = 1200 trials.
-For each trial, we had 5 to 7 attemps depending on the error rate, which means a total of 6,000 to 8,400 total samples.
-After the experiment, we actually recorded 6963 samples.
+Five to seven attempts were recorded per trial, depending on error rate, and 6963 samples were recorded in total.
 
@@
 % =============================================================================
-\section{Result}
+\section{Results}
 % =============================================================================
-The dependent variables are the success rate, the time taken per trial and the trace data when available.
-This allowed us to compute the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in~\cite{Markussen2014}, according to the following formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.
+The dependent variables we analyse are the success rate, the time taken per trial, and the trace data when available.
+From these, we computed the dependent measure error rate, defined as the percentage of unsucessful attempts, and the text entry rate, measured in Words Per Minute (WPM) as in~\cite{Markussen2014}\footnote{$WPM = |T|/s \times 60/5$, where $|T|$ is the length of the transcribed string, s is time in seconds}.
 
+\subsection{Word Error Rate}
 We found one input word (\cdt{lay}) presented some very unusual behavior.
-Its error rate was at $89.5\%$ while the \cdt{WORD} mean is $20.1\%$ and no other word had an error rate higher than $30\%$.
-The reason is that the recogniser promotes words of higher prior probability in the language model, \textit{``Larry''}, \textit{``last''} or \textit{``Katy''}.
-For the all subsequent analysis of error rate, \cdt{lay} is removed from the dataset.
+Its error rate was at $89.5\%$ while the \cdt{WORD} mean was $20.1\%$ and no other word had an error rate higher than $30\%$.
+The reason was that the recogniser promoted words of higher prior probability in the language model, \textit{``Larry''}, \textit{``last''} or \textit{``Katy''}.
+For all subsequent analysis of error rate, \cdt{lay} was removed from the dataset.
 
@@
 
-A statistical analysis showed a significant main effect of \cdt{DEVICE} ($F_{1,11} = 90.15, p < 0.0001$) on input rate, see \autoref{fig:wpm_DEVICE_SHAPE}, with mean values for \cdt{TP1} and \cdt{OP1} equal to respectively $29.2 WPM$ and $13.7 WPM$.
+\subsection{Word Input Rate}
+A statistical analysis showed a significant main effect of \cdt{DEVICE} ($F_{1,11} = 90.15, p < 0.0001$) on input rate, see \autoref{fig:wpm_DEVICE_SHAPE}, with mean values for \cdt{TP1} and \cdt{OP1} equal to $29.2 WPM$ and $13.7 WPM$, respectively.
 We also looked at pairwise comparison\footnote{Post-hoc analysis were adjusted using Bonferroni correction.} for \cdt{OPTICAL} and could not find a statistical difference in the mean input rate.
@@
 
-\subsection{Effect of orientation}
-The design of the experiment also includes \cdt{ORIENTATION}.
-This condition has so far been excluded from the SIZE analysis and can be difficult to apprehend since not only the \cdt{ORIENTATION} of the visual feedback is different but its size also.
-The main result is the average pointer speed in display space at $303.75 pixel/s$, higher than the PORTRAIT orientation at an average $228 pixel/s$.
-It is important to keep in mind that the keyboard area in LANDSCAPE is $61cm^2$, bigger than PORTRAIT at $44cm^2$, which can explain the faster pointer speed but is not responsible for a higher input rate.
+\subsection{Effect of Orientation}
+The design of the experiment also included \cdt{ORIENTATION}.
+This condition has so far been excluded from the SIZE analysis and can be difficult to apprehend since not only the \cdt{ORIENTATION} of the visual feedback changes, but also its size.
+The main result is that the average pointer speed in display space at $303.75$ pixel/s was higher than the PORTRAIT orientation at an average $228$ pixel/s.
+It is important to note that the keyboard area in LANDSCAPE ($61$ $cm^2$) is bigger than PORTRAIT ($44$ $cm^2$), which can explain the faster pointer speed, but is not responsible for a higher input rate.
 
-\subsection{Qualitative data}
-At the end of the experiment, participants were invited to provide some informal feedback, rank the different level in order of preference and fill in the NASA-TLX form, the tablet interaction was ranked best by all the participants except one.
+\subsection{Qualitative Data}
+At the end of the experiment, participants were invited to provide some informal feedback, rank the different level in order of preference and fill in the NASA-TLX form.
+The tablet interaction was ranked best by all participants except one.
 \cdt{OP4} was consistently ranked last, while \cdt{OP2} and \cdt{OL2} had equal ranking in second position.
@@
 
-From the participants feedback, it shows that \cdt{OP1} was a demanding interaction.
-It also showed that increasing the surface size to the extent of \cdt{OP4} is physically demanding.
-Recommended sizes for interaction will thus be \cdt{OP2} or \cdt{OL2} (note that the performance is also dependent on the $CD_{gain}$).
-Moreover, some participants reported the texture having an impact of the interaction.
-Levesque et al.~\cite{Levesque2011} have used friction in a dynamic manner to improve target acquisition.
+Participant feedback showed that \cdt{OP1} was a demanding interaction, and that increasing the surface size to the extent of \cdt{OP4} was also physically demanding.
+We would therefore recommended sizes \cdt{OP2} or \cdt{OL2} for interaction, noting that performance is also dependent on $CD_{gain}$.
+Some participants reported texture having an impact on the interaction.
+Levesque et al.~\cite{Levesque2011} used friction in a dynamic manner to improve target acquisition.
 Given the relatively low performance of the participants, it could be interesting to explore if tactile feedback could help controlling an indirect pointer.
@@
 We designed an experiment with two research goals in mind.
-First, we compared the typing performance of our system against a control condition which was a direct interaction on a tablet.
+First, we compared the typing performance of our optical indirect system with a control condition with direct interaction on a tablet.
 Second, we studied the influence of size and aspect ratio of the input surface on gesture typing performance.
-We showed that the participants could enter text at half the input rate and the same error rate on surface at least twice the size of the current visual feedback.
-We also showed that the input rate is largely independent of the surface size across the range of sizes we were able to examine.
-Gesture typing has promise for interaction outside of mobile devices (e.g.
-for motor impaired users who struggle with capacitive touch technologies).
+We showed that the participants could enter text at half the input rate and the same error rate on a surface at least twice the size of the visual feedback.
+We also showed that the input rate is largely independent of surface size across the range of sizes we examined.
+Gesture typing has promise for interaction outside of mobile devices, e.g.
+for motor impaired users who struggle with capacitive touch technologies.
 This paper indicates that while camera-tracked gesture typing performance is usable, input rates are lower than touchscreen performance, and this is not influenced by scaling of the input.

### 2017-08-22 paper.tex
before merging overleaf

--- 2017-08-22 paper.tex
@@
 
-\title{Gesture Typing on Virtual Tabletop: Effect of Input on Performance}
+\title{Gesture Typing on Virtual Tabletop: Effect of Input Dimensions on Performance}
 
@@
 % Paper metadata (use plain text, for PDF inclusion and later re-using, if desired)
-\def\plaintitle{Gesture Typing on Virtual Surfaces: Effect of Input on Performance}
+\def\plaintitle{Gesture Typing on Virtual Surfaces: Effect of Input Dimensions on Performance}
 \def\plainauthor{Name}
-\def\plainkeywords{Gesture Input, Gesture Keyboard, Mobile, Continuous Interaction, Tabletop, Text Input}
+\def\plainkeywords{Indirect Interaction, Gesture Input, Gesture Keyboard, Mobile, Continuous Interaction, Tabletop, Text Input}
 \def\plaingeneralterms{Documentation, Standardization}
@@
 \begin{abstract}
-The association of tabletop interaction with gesture typing presents potential for physically impaired users.
+The association of tabletop interaction with gesture typing presents interaction potential, especially for situationally impaired or physically impaired users.
 In this work, we use depth cameras to create touch surfaces on regular tabletops.
@@
 
-Depth cameras~\cite{Harrison2011,Xiao2016,Sridhar2017} have usually been employed to create such surfaces, while other work have tried to combine different sensor sources~\cite{Wen2016}.
-However, the touch classification, critical to the quality of the interaction remains problematic (sub capacitive quality?) and few systems are available for all camera types.
+Depth cameras~\cite{Harrison2011,Sridhar2017,Xiao2016} have usually been employed to create such surfaces, while other work have tried to combine different sensor sources~\cite{Wen2016}.
+However, the touch classification, critical to the quality of the interaction remains challenging~\cite{Xiao2016} and is traditionally addressed by hand-tuning parameters and thresholding.
 
@@
 
-% \textcolor{red}{potential for more space}
-
-% \textcolor{red}{potential for more space}
-
 In this context, the task of gesture typing~\cite{Kristensson2004} is interesting to research for the following reasons.
-Text-input remains a major activity~\cite{McGregor2014} on mobile device occupying up to 40\% of the user interaction time, the technique lends itself well to optical systems by limiting the requirement for repetitive target acquisition or touch classification, and little is known about the impact of input size (one of the free parameters of virtual surface) on writing performance.
+Text-input is a major activity~\cite{McGregor2014} on mobile device taking up to 40\% of the user interaction time, the technique lends itself well to optical systems by limiting the requirement for repetitive target acquisition or touch classification, and little is known about the impact of input size (one of the free parameters of virtual surfaces) on writing performance as other works are focused on pointing~\cite{Casiez2008,Gilliot2014} or bimanual task\cite{Schmidt2009}.
 
@@
 % =============================================================================
-The envisioned interaction scenario is shown in ~\autoref{fig:banner} with its implementation in \autoref{fig:system_insitu_0}.
+The envisioned interaction scenario is shown in ~\autoref{fig:banner}.
+A camera overlooks the interaction surface and performs the hand and finger tracking, as well as the touch detection.
+The system mapped in an indirect manner the user's pointer from the virtual surface onto the tablet screen.
+A 3-state button model is used with visual feedback indicating hover positions and continuous touches.
+An audio cue indicates whenever a new touch down event is detected.
+
+\section{Touch classification}
+
+ % with its implementation in \autoref{fig:system_insitu_0}.
 We use an Intel Realsense SR300 obliquely mounted on a tripod that overlooks the interaction surface.
@@
 
-The pointer tracked by the system is defined as the point cloud closest to the camera in the depth direction ($y$ on ~\autoref{fig:pointcloud}); there is no explicit finger or hand modeling performed.
-Instead, the assumption is that the interacting finger is the furthest protruding object from the user.
+% The touch classification problem is usually solved via flood-filling~\cite{Harrison2011,Sridhar2017,Xiao2016}. However, it
+This task is well suited to a supervised learning approach provided availability of a training dataset.
+One of the authors video-taped 6 minutes of interactions creating 3 distincts datasets for 3 different fingers (index, thumb and pinky) totalling 10800 frames.
+The datasets are balanced, including half touching frames and half hovering frames with some frames out of range and 2 camera angles.
+The features are 20 bins histogram of z-values of the detected fingertip pointcloud, in red on \autoref{fig:pointcloud}, which provides some orientation and shape invariance.
 
-The touch classification problem is usually solved via flood-filling~\cite{Harrison2011,Xiao2016,Sridhar2017}.
-However, it is well suited to a supervised learning approach provided availability of a training dataset.
-One of the authors video-taped 6 minutes of interactions, half touching and half hovering with some frames out of range while using the index, thumb and pinky at close and long range.
-From the fingertip pointcloud, we use the histogram of z-values as features (providing some orientation and shape invariance) and a neural network classifier.
-We tested the generalisation with different cross validation splits and obtained an averaged 0.96 AUC for the ROC, see~\autoref{fig:roc_auc}.
-The ROC curves led us to choose an operating point of 0.5 for the live system, which we trained on the whole dataset for 75 epochs.
-
+This is fed to a binary neural network classifier: 20 inputs dimensions, 2 layers of 64 units with Relu activation and $0.5$ dropout, and a fully connected layer with sigmoid activation.
+The loss is set to binary cross-entropy and rmsprop reguliser.
+We train this model over 50 epochs and cross validated with 3 fold validation (different fingers are validated against each others).
+The generalisation over different fingers is satisfying, as well as the performance averaging 0.96 AUC for the ROC, see~\autoref{fig:roc_auc}.
+The model used for the user study was trained over the full dataset over 75 epochs.
+As an improvment, it is worth noting the possibility to delegate the feature extraction step to a model capable of inferring features directly, such as a convolutional neural network.
 % \begin{figure}
@@
 Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance (Accot et al.~\cite{Accot2001} demonstrated a U-shaped curve in performance against input scale).
-Our independent variable was thus the control surface dimensions, which is effectively changing the control-display gain ($CD_{gain}$) defined as the ratio between $V_{display}$ and $V_{control}$.
+Our independent variable was thus the control surface dimensions, which is effectively changing the control-display gain ($CD_{gain}$) defined as in \cite{Casiez2008} by the the ratio of the pointer velocity to device velocity.
+
+We gathered 12 participants, all right handed, without requirement on gesture typing experience.
+A repeated measures within-subjects design was used.
+There were 3 conditions (\cdt{device}, \cdt{size} and \cdt{orientation}) and 5 level combinations were evaluated.
+We adopted a 3-symbols naming convention: the first letter represents the DEVICE, the second marks the ORIENTATION and the number represents the scaling factor as SIZE.
+5 level combinations were evaluated, see \autoref{tab:cdt} below for details.
+
+% A repeated measures within-subjects design was used.
+There were 3 conditions and 5 level combinations were evaluated, see \autoref{tab:cdt} below for details.
+The presentation order of the 5 combinations was randomised and presented a uniform distribution across the participants.
 
@@
 
-We gathered 12 participants, all right handed, without requirement on gesture typing experience.
-A repeated measures within-subjects design was used.
-There were 3 conditions and 5 level combinations were evaluated, see \autoref{tab:cdt} below for details.
-The presentation order of the 5 combinations was randomised and presented a uniform distribution across the participants.
+The apparatus, see \autoref{fig:system_insitu_0}, implements the system description above.
+An Intel Realsense camera is used, and the processing is performed on a desktop computer.
+An Android tablet running custom software produces the audiovisual feedback.
+The pointer tracked by the system is defined as the point cloud closest to the camera in the depth direction ($y$ on ~\autoref{fig:pointcloud}); there is no explicit finger or hand modeling performed.
+Instead, the assumption is that the interacting finger is the furthest protruding object from the user.
 
@@
 
+\subsection{Effect on error rate}
 A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77, p < 0.0001$) on error rate with mean value for \cdt{TP1} and \cdt{OP1} equal to respectively 6.2\% and 26.1\%, see \autoref{fig:err_DEVICE_SHAPE}.
@@
 
+\subsection{Effect on input rate}
 A statistical analysis showed a significant main effect of \cdt{DEVICE} ($F_{1,11} = 90.15, p < 0.0001$) on input rate, see \autoref{fig:wpm_DEVICE_SHAPE}, with mean values for \cdt{TP1} and \cdt{OP1} equal to respectively $29.2 WPM$ and $13.7 WPM$.
@@
 
-We proposed a machine learning approach to fingertip touch detection.
+% We proposed a machine learning approach to fingertip touch detection.
 This problem is traditionally addressed by hand-tuning parameters and thresholding.

### 2017-08-22 paper.tex

@@
 The combination of indirect optically tracked input and potentially projected display (“virtual surfaces”) has several interesting properties.
-Virtual surfaces, as opposed to touch screens, are well suited to tackle the palm rejection problem and the occlusion problem, offer a choice of size, aspect ratio and texture of the input space and allow interactions with dirty or wet hands.
+Virtual surfaces, as opposed to touch screens, are well suited to tackle the palm rejection problem and the occlusion problem, offer a choice of size, aspect ratio and texture of the input space, and allow interactions with dirty or wet hands.
 
-Depth cameras~\cite{Harrison2011,Sridhar2017,Xiao2016} have usually been employed to create such surfaces, while other work have tried to combine different sensor sources~\cite{Wen2016}.
+Depth cameras~\cite{Harrison2011,Sridhar2017,Xiao2016} have usually been employed to create such surfaces, while other work have combined different sensor sources~\cite{Wen2016}.
 However, the touch classification, critical to the quality of the interaction remains challenging~\cite{Xiao2016} and is traditionally addressed by hand-tuning parameters and thresholding.
@@
 
-In this context, the task of gesture typing~\cite{Kristensson2004} is interesting to research for the following reasons.
-Text-input is a major activity~\cite{McGregor2014} on mobile device taking up to 40\% of the user interaction time, the technique lends itself well to optical systems by limiting the requirement for repetitive target acquisition or touch classification, and little is known about the impact of input size (one of the free parameters of virtual surfaces) on writing performance as other works are focused on pointing~\cite{Casiez2008,Gilliot2014} or bimanual task\cite{Schmidt2009}.
+In this context, the task of gesture typing~\cite{Kristensson2004} is an interesting research focus.
+Text-input is a major activity~\cite{McGregor2014} on mobile device taking up to 40\% of the user interaction time.
+The technique lends itself well to optical systems by limiting the requirement for repetitive target acquisition or touch classification.
+Yet, little is known about the impact of input size (one of the free parameters of virtual surfaces) on writing performance as other works focused on pointing~\cite{Casiez2008,Gilliot2014} or bimanual task\cite{Schmidt2009}.
 
-This work specifically focus on two goals: to report on a supervised learning approach to the fingertip touch classification task and to study the influence of input space (size and aspect ratio) on gesture typing performance.
+This work specifically focuses on two goals: to report on a supervised learning approach to the fingertip touch classification task and to study the influence of input space (size and aspect ratio) on gesture typing performance using virtual tabletops.
 % \begin{itemize}
@@
 
-\section{Touch classification}
+\section{Touch Classification}
 
@@
 % =============================================================================
-\section{User study}
+\section{User Study}
 % =============================================================================
@@
 First we wanted to evaluate the usability of our system, for this purpose we included a condition with an interaction directly on the tablet.
-Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance (Accot et al.~\cite{Accot2001} demonstrated a U-shaped curve in performance against input scale).
-Our independent variable was thus the control surface dimensions, which is effectively changing the control-display gain ($CD_{gain}$) defined as in \cite{Casiez2008} by the the ratio of the pointer velocity to device velocity.
+Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance.
+Accot et al.~\cite{Accot2001} demonstrated a U-shaped curve in performance against input scale.
+Our independent variable were thus the control surface dimensions, which effectively change the control-display gain ($CD_{gain}$) defined as in \cite{Casiez2008} by the the ratio of the pointer velocity to device velocity.
 
-We gathered 12 participants, all right handed, without requirement on gesture typing experience.
+We recruited 12 participants, all right handed, without requirement on gesture typing experience.
 A repeated measures within-subjects design was used.
@@
 % =============================================================================
-\section{Result}
+\section{Results}
 % =============================================================================
-The dependent variables are the success rate, the time taken per trial and the trace data when available.
-This allowed us to compute the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in~\cite{Markussen2014}, according to the following formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.
+The dependent variables we analyse are the success rate, the time taken per trial and the trace data when available.
+From these, we computed the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in~\cite{Markussen2014}, according to the following formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.
 
 We found one input word (\cdt{lay}) presented some very unusual behavior.
-Its error rate was at $89.5\%$ while the \cdt{WORD} mean is $20.1\%$ and no other word had an error rate higher than $30\%$.
-The reason is that the recogniser promotes words of higher prior probability in the language model, \textit{``Larry''}, \textit{``last''} or \textit{``Katy''}.
-For the all subsequent analysis of error rate, \cdt{lay} is removed from the dataset.
+Its error rate was at $89.5\%$ while the \cdt{WORD} mean was $20.1\%$ and no other word had an error rate higher than $30\%$.
+The reason was that the recogniser promoted words of higher prior probability in the language model, \textit{``Larry''}, \textit{``last''} or \textit{``Katy''}.
+For the all subsequent analysis of error rate, \cdt{lay} was removed from the dataset.
 
 \subsection{Effect on error rate}
-A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77, p < 0.0001$) on error rate with mean value for \cdt{TP1} and \cdt{OP1} equal to respectively 6.2\% and 26.1\%, see \autoref{fig:err_DEVICE_SHAPE}.
+A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77, p < 0.0001$) on error rate with mean value for \cdt{TP1} and \cdt{OP1} equal to 6.2\% and 26.1\%, respectively, see \autoref{fig:err_DEVICE_SHAPE}.
 A statistical analysis also showed a significant main effect of \cdt{SIZE} ($F_{2,22} = 10.99, p < 0.001$) on the error rate.
@@
 \subsection{Effect on input rate}
-A statistical analysis showed a significant main effect of \cdt{DEVICE} ($F_{1,11} = 90.15, p < 0.0001$) on input rate, see \autoref{fig:wpm_DEVICE_SHAPE}, with mean values for \cdt{TP1} and \cdt{OP1} equal to respectively $29.2 WPM$ and $13.7 WPM$.
+A statistical analysis showed a significant main effect of \cdt{DEVICE} ($F_{1,11} = 90.15, p < 0.0001$) on input rate, see \autoref{fig:wpm_DEVICE_SHAPE}, with mean values for \cdt{TP1} and \cdt{OP1} equal to $29.2 WPM$ and $13.7 WPM$, respectively.
 We also looked at pairwise comparison\footnote{Post-hoc analysis were adjusted using Bonferroni correction.} for \cdt{OPTICAL} and could not find a statistical difference in the mean input rate.
@@
 The design of the experiment also includes \cdt{ORIENTATION}.
-This condition has so far been excluded from the SIZE analysis and can be difficult to apprehend since not only the \cdt{ORIENTATION} of the visual feedback is different but its size also.
+This condition has so far been excluded from the SIZE analysis and can be difficult to apprehend since not only the \cdt{ORIENTATION} of the visual feedback changes, but its size also.
 The main result is the average pointer speed in display space at $303.75 pixel/s$, higher than the PORTRAIT orientation at an average $228 pixel/s$.
-It is important to keep in mind that the keyboard area in LANDSCAPE is $61cm^2$, bigger than PORTRAIT at $44cm^2$, which can explain the faster pointer speed but is not responsible for a higher input rate.
+It is important to keep in mind that the keyboard area in LANDSCAPE ($61cm^2$), bigger than PORTRAIT ($44cm^2$), which can explain the faster pointer speed but is not responsible for a higher input rate.
 
 \subsection{Qualitative data}
-At the end of the experiment, participants were invited to provide some informal feedback, rank the different level in order of preference and fill in the NASA-TLX form, the tablet interaction was ranked best by all the participants except one.
+At the end of the experiment, participants were invited to provide some informal feedback, rank the different level in order of preference and fill in the NASA-TLX form, the tablet interaction was ranked best by all participants except one.
 \cdt{OP4} was consistently ranked last, while \cdt{OP2} and \cdt{OL2} had equal ranking in second position.
@@
 
-From the participants feedback, it shows that \cdt{OP1} was a demanding interaction.
-It also showed that increasing the surface size to the extent of \cdt{OP4} is physically demanding.
-Recommended sizes for interaction will thus be \cdt{OP2} or \cdt{OL2} (note that the performance is also dependent on the $CD_{gain}$).
-Moreover, some participants reported the texture having an impact of the interaction.
+Participants feedback, it shows that \cdt{OP1} was a demanding interaction, and that increasing the surface size to the extent of \cdt{OP4} is physically demanding.
+We would therefore recommend sizes \cdt{OP2} or \cdt{OL2} for interaction, noting that the performance is also dependent on the $CD_{gain}$).
+Some participants reported the texture having an impact of the interaction.
 Levesque et al.~\cite{Levesque2011} have used friction in a dynamic manner to improve target acquisition.
@@
 We designed an experiment with two research goals in mind.
-First, we compared the typing performance of our system against a control condition which was a direct interaction on a tablet.
+First, we compared the typing performance of our optical indirect system against a control condition which was a direct interaction on a tablet.
 Second, we studied the influence of size and aspect ratio of the input surface on gesture typing performance.
-We showed that the participants could enter text at half the input rate and the same error rate on surface at least twice the size of the current visual feedback.
-We also showed that the input rate is largely independent of the surface size across the range of sizes we were able to examine.
-Gesture typing has promise for interaction outside of mobile devices (e.g.
-for motor impaired users who struggle with capacitive touch technologies).
+We showed that the participants could enter text at half the input rate and the same error rate on surface at least twice the size of the visual feedback.
+We also showed that the input rate is largely independent of surface size across the range of sizes we were able to examine.
+Gesture typing has promise for interaction outside of mobile devices, e.g.
+for motor impaired users who struggle with capacitive touch technologies.
 This paper indicates that while camera-tracked gesture typing performance is usable, input rates are lower than touchscreen performance, and this is not influenced by scaling of the input.

### 2017-08-22 paper.tex
final submitted

@@
 \begin{abstract}
-The association of tabletop interaction with gesture typing presents interaction potential, especially for situationally impaired or physically impaired users.
+The association of tabletop interaction with gesture typing presents interaction potential for situationally or physically impaired users.
 In this work, we use depth cameras to create touch surfaces on regular tabletops.
@@
 We follow with a gesture typing study that compares our system with a control tablet scenario and explore the influence of input size and aspect ratio of the virtual surface on the text input performance.
-We show that novice users perform with the same error rate at half the input rate with our system as compared to the control condition, that an input size between A5 and A4 present the best tradeoff between performance and user preference and that user's indirect tracking ability seems to be the overall performance limiting factor.
-
-\textcolor{red}{potential for more space}
-
+We show that novice users perform with the same error rate at half the input rate with our system as compared to the control condition, that an input size between A5 and A4 present the best tradeoff between performance and user preference and that users' indirect tracking ability seems to be the overall performance limiting factor.
 \end{abstract}
@@
 The technique lends itself well to optical systems by limiting the requirement for repetitive target acquisition or touch classification.
-Yet, little is known about the impact of input size (one of the free parameters of virtual surfaces) on writing performance as other works focused on pointing~\cite{Casiez2008,Gilliot2014} or bimanual task\cite{Schmidt2009}.
+Yet, little is known about the impact of input size (one of the free parameters of virtual surfaces) on writing performance as other works focused on pointing~\cite{Casiez2008,Gilliot2014} or bimanual task~\cite{Schmidt2009}.
 
@@
 The envisioned interaction scenario is shown in ~\autoref{fig:banner}.
-A camera overlooks the interaction surface and performs the hand and finger tracking, as well as the touch detection.
-The system mapped in an indirect manner the user's pointer from the virtual surface onto the tablet screen.
+A camera overlooks the interaction surface and a computing device performs the hand and fingers tracking as well as the touch detection.
+The system map in an indirect manner the user's pointer from the virtual surface onto the device's screen.
 A 3-state button model is used with visual feedback indicating hover positions and continuous touches.
-An audio cue indicates whenever a new touch down event is detected.
+Audio cues indicate whenever a new touch down event is detected.
 
 \section{Touch Classification}
+This task is well suited to a supervised learning approach provided a training dataset is available.
+One of the authors video-taped 6 minutes of interactions creating 3 distincts datasets for 3 different fingers (index, thumb and pinky) totalling 10800 frames.
+The datasets are balanced with regards to touching frames and hovering frames, a few out-of-range frames are also included.
+The classification features are 20-bins histograms of z-values of the detected fingertip pointcloud, in red on~\autoref{fig:pointcloud}, which provides some orientation and shape invariance.
 
- % with its implementation in \autoref{fig:system_insitu_0}.
-We use an Intel Realsense SR300 obliquely mounted on a tripod that overlooks the interaction surface.
-Finger detection and tracking is performed on a desktop computer.
-The system implements the 3 state button model, with its touch events sent to the Android tablet running a custom application overwriding its input event system.
-In addition, the tablet displays hover state as a red marker at the pointer’s position, touching is displayed through the Android debugging facility as a cross spanning the screen and continuous touch is displayed as a segment line.
-An audio feedback is produced on each touch down event.
+This is fed to a binary neural network classifier: 20 inputs dimensions, 2 layers of 64 units with Relu activation and $0.5$ dropout, and a fully connected layer with sigmoid activation.
+The loss is set to binary cross-entropy and the optimiser to rmsprop.
+We train the model over 50 epochs and cross-validate with 3 folds (different fingers are validated against each others).
+The generalisation over different fingers is satisfying as shown by the average performance of 0.96 AUC for the ROC,
+see~\autoref{fig:roc_auc}.
+The model used for the user study was trained over the full dataset over 75 epochs.
+As an improvment, it is worth noting the possibility to delegate the feature extraction step to a model capable of inferring features directly, such as a convolutional neural network.
 
-% The touch classification problem is usually solved via flood-filling~\cite{Harrison2011,Sridhar2017,Xiao2016}. However, it
-This task is well suited to a supervised learning approach provided availability of a training dataset.
+% The task is well suited to a supervised learning approach provided availability of a training dataset.
 One of the authors video-taped 6 minutes of interactions creating 3 distincts datasets for 3 different fingers (index, thumb and pinky) totalling 10800 frames.
@@
 
-This is fed to a binary neural network classifier: 20 inputs dimensions, 2 layers of 64 units with Relu activation and $0.5$ dropout, and a fully connected layer with sigmoid activation.
+% This is fed to a binary neural network classifier: 20 inputs dimensions, 2 layers of 64 units with Relu activation and $0.5$ dropout, and a fully connected layer with sigmoid activation.
 The loss is set to binary cross-entropy and rmsprop reguliser.
@@
 As an improvment, it is worth noting the possibility to delegate the feature extraction step to a model capable of inferring features directly, such as a convolutional neural network.
+
 % \begin{figure}
@@
 
-We recruited 12 participants, all right handed, without requirement on gesture typing experience.
-A repeated measures within-subjects design was used.
-There were 3 conditions (\cdt{device}, \cdt{size} and \cdt{orientation}) and 5 level combinations were evaluated.
-We adopted a 3-symbols naming convention: the first letter represents the DEVICE, the second marks the ORIENTATION and the number represents the scaling factor as SIZE.
-5 level combinations were evaluated, see \autoref{tab:cdt} below for details.
-
-% A repeated measures within-subjects design was used.
-There were 3 conditions and 5 level combinations were evaluated, see \autoref{tab:cdt} below for details.
-The presentation order of the 5 combinations was randomised and presented a uniform distribution across the participants.
-
 \marginpar{
@@
 }
+
+We recruited 12 participants, all right handed, without requirement on gesture typing experience.
+A repeated measures within-subjects design was used.
+There were 3 conditions (\cdt{device}, \cdt{size} and \cdt{orientation}) and 5 level combinations were evaluated, see \autoref{tab:cdt} below for details.
+We adopted a 3-symbols naming convention: the first letter represents the DEVICE, the second marks the ORIENTATION and the number represents the scaling factor as SIZE.
 
@@
 The main result is the average pointer speed in display space at $303.75 pixel/s$, higher than the PORTRAIT orientation at an average $228 pixel/s$.
-It is important to keep in mind that the keyboard area in LANDSCAPE ($61cm^2$), bigger than PORTRAIT ($44cm^2$), which can explain the faster pointer speed but is not responsible for a higher input rate.
+It is important to keep in mind that the keyboard area in LANDSCAPE ($61cm^2$), bigger than PORTRAIT ($44cm^2$), which can explain the faster pointer speed but is not responsible for a higher input rate due to longer traces to produces.
 
@@
 Because participants are capable of adapting their motor behavior and their control bahavior, another explanation for the upper bound in text input is that the indirect nature of the interaction is the limiting factor.
-(more research in that direction?)
 
-Participants feedback, it shows that \cdt{OP1} was a demanding interaction, and that increasing the surface size to the extent of \cdt{OP4} is physically demanding.
+Participants feedback showed that \cdt{OP1} was a demanding interaction, and that increasing the surface size to the extent of \cdt{OP4} was also physically demanding.
 We would therefore recommend sizes \cdt{OP2} or \cdt{OL2} for interaction, noting that the performance is also dependent on the $CD_{gain}$).

### 2017-08-28 paper.tex
add acknowledgements, minor changes to touch classification

@@
 %% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
-% \copyrightinfo{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
+\copyrightinfo{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
 Copyrights for components of this work owned by others than ACM must be honored.
@@
 \\
-% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
-% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
-% DOI string from ACM form confirmation}
+{\emph{ISS '17}}, October 17--20, 2017, Brighton, United Kingdom. \\
+Copyright \copyright~2017 ACM ISBN 978-1-4503-4691-7/17/10. \\
+10.1145/3132272.3135074}
 %% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
+
+% \CopyrightYear{2017}
+% \setcopyright{rightsretained}
+% \conferenceinfo{ISS '17}{October 17--20, 2017, Brighton, United Kingdom}\isbn{978-1-4503-4691-7/17/10}
+% \doi{https://doi.org/10.1145/3132272.3135074}
 
@@
 This work specifically focuses on two goals: to report on a supervised learning approach to the fingertip touch classification task and to study the influence of input space (size and aspect ratio) on gesture typing performance using virtual tabletops.
-% \begin{itemize}
-% \item to report on a supervised learning approach to the fingertip touch classification task.
-% \item to study the influence of input space (size and aspect ratio) on gesture typing performance.
-% \end{itemize}
 
@@
 One of the authors video-taped 6 minutes of interactions creating 3 distincts datasets for 3 different fingers (index, thumb and pinky) totalling 10800 frames.
-The datasets are balanced with regards to touching frames and hovering frames, a few out-of-range frames are also included.
-The classification features are 20-bins histograms of z-values of the detected fingertip pointcloud, in red on~\autoref{fig:pointcloud}, which provides some orientation and shape invariance.
+The datasets are balanced with regards to touching frames and hovering frames and include a minor portion of out-of-range frames.
+The classification features are chosen as 20-bins histograms of z-values of the detected fingertip pointcloud, in red on~\autoref{fig:pointcloud}, which provides some orientation and shape invariance.
 
 This is fed to a binary neural network classifier: 20 inputs dimensions, 2 layers of 64 units with Relu activation and $0.5$ dropout, and a fully connected layer with sigmoid activation.
-The loss is set to binary cross-entropy and the optimiser to rmsprop.
-We train the model over 50 epochs and cross-validate with 3 folds (different fingers are validated against each others).
-The generalisation over different fingers is satisfying as shown by the average performance of 0.96 AUC for the ROC,
-see~\autoref{fig:roc_auc}.
+We use the binary cross-entropy as loss function and rmsprop for the optimiser.
+We train the model over 50 epochs and cross-validate with 3 folds (different fingers are thus validated against each others).
+The generalisation over different fingers is satisfying as shown by the average performance of 0.96 AUC for the ROC, see~\autoref{fig:roc_auc}.
 The model used for the user study was trained over the full dataset over 75 epochs.
 As an improvment, it is worth noting the possibility to delegate the feature extraction step to a model capable of inferring features directly, such as a convolutional neural network.
-
-% The task is well suited to a supervised learning approach provided availability of a training dataset.
-One of the authors video-taped 6 minutes of interactions creating 3 distincts datasets for 3 different fingers (index, thumb and pinky) totalling 10800 frames.
-The datasets are balanced, including half touching frames and half hovering frames with some frames out of range and 2 camera angles.
-The features are 20 bins histogram of z-values of the detected fingertip pointcloud, in red on \autoref{fig:pointcloud}, which provides some orientation and shape invariance.
-
-% This is fed to a binary neural network classifier: 20 inputs dimensions, 2 layers of 64 units with Relu activation and $0.5$ dropout, and a fully connected layer with sigmoid activation.
-The loss is set to binary cross-entropy and rmsprop reguliser.
-We train this model over 50 epochs and cross validated with 3 fold validation (different fingers are validated against each others).
-The generalisation over different fingers is satisfying, as well as the performance averaging 0.96 AUC for the ROC, see~\autoref{fig:roc_auc}.
-The model used for the user study was trained over the full dataset over 75 epochs.
-As an improvment, it is worth noting the possibility to delegate the feature extraction step to a model capable of inferring features directly, such as a convolutional neural network.
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=\linewidth]{./figures/pointcloud.pdf}
-%   \caption{Insert a caption below each figure.}
-%   \label{fig:sample}
-% \end{figure}
 
@@
 
-% \section{Acknowledgements}
-% We would like to thank the EU Moregrasp project for funding the research, as well as the patients in the spinal unit in Heidelberg for providing early feedback about the interaction technique.
-
-% \newpage
+\section{Acknowledgements}
+We thank the European Commission and Horizon 2020 for funding this research (award number 643955).
 

### 2017-08-28 paper.tex
caption for touch classification
small edits

@@
 One of the authors video-taped 6 minutes of interactions creating 3 distincts datasets for 3 different fingers (index, thumb and pinky) totalling 10800 frames.
-The datasets are balanced with regards to touching frames and hovering frames and include a minor portion of out-of-range frames.
+The datasets are balanced with regards to touching frames and hovering frames and include a minor fraction of out-of-range frames.
+
 The classification features are chosen as 20-bins histograms of z-values of the detected fingertip pointcloud, in red on~\autoref{fig:pointcloud}, which provides some orientation and shape invariance.
+This is fed to a binary neural network classifier: 20 inputs dimensions, 2 layers of 64 hidden units with rectified linear (ReLU) activation and 50\% dropout, and a fully connected output layer with sigmoid activation.
+We use the binary cross-entropy as loss function and rmsprop for the optimiser.
+Finally, we train the model over 50 epochs and perform a 3 folds cross-validation across the 3 datasets, different fingers are thus validated against each others.
+This provides an estimate for the generalisation power of the model when classifying unseen users as shown by the averaged 0.96 AUC for the ROC, see~\autoref{fig:roc_auc}.
 
-This is fed to a binary neural network classifier: 20 inputs dimensions, 2 layers of 64 units with Relu activation and $0.5$ dropout, and a fully connected layer with sigmoid activation.
-We use the binary cross-entropy as loss function and rmsprop for the optimiser.
-We train the model over 50 epochs and cross-validate with 3 folds (different fingers are thus validated against each others).
-The generalisation over different fingers is satisfying as shown by the average performance of 0.96 AUC for the ROC, see~\autoref{fig:roc_auc}.
 The model used for the user study was trained over the full dataset over 75 epochs.
@@
     \includegraphics[width=\linewidth]{figures/roc_auc.pdf}
-    \caption{Performance of the fingertip touch classification.}~\label{fig:roc_auc}
+    \caption{Performance of the fingertip touch classification with 3-fold cross validation.
+Each split trains on two finger types (among three) and validates against the remaining one.}~\label{fig:roc_auc}
 \end{figure}
@@
 The dependent variables we analyse are the success rate, the time taken per trial and the trace data when available.
-From these, we computed the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in~\cite{Markussen2014}, according to the following formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.
+From these, we computed the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in~\cite{Markussen2014}, according to the following formula: $WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string and $s$ is time in seconds.
 
@@
 
-% We proposed a machine learning approach to fingertip touch detection.
-This problem is traditionally addressed by hand-tuning parameters and thresholding.
-In contrast, we showed that using a machine learning approach can solve the problem and generalise well enough to be used in an experiment with 12 unseen participants.
-One improvement could be to delegate the feature extraction step (an histogram of z-values in our case) to a model capable of inferring features directly, such as a convolutional neural network.
-Since the data is cheap to record, we could see this approach being used for many specific cases.
-
 We conducted a user study investigating the usability of the system and showed that for control dimension at least twice the display dimension novice users were capable the same error rate at half the input rate.

### 2017-08-28 paper.tex
copyright, keywords

@@
 %% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
-\copyrightinfo{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
-Copyrights for components of this work owned by others than ACM must be honored.
-Abstracting with credit is permitted.
-To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee.
-Request permissions from permissions@acm.org.
+\copyrightinfo{Permission to make digital or hard copies of part or all of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
+Copyrights for third-party components of this work must be honored.
+For all other uses, contact the Owner/Author.
 \\
-{\emph{ISS '17}}, October 17--20, 2017, Brighton, United Kingdom. \\
-Copyright \copyright~2017 ACM ISBN 978-1-4503-4691-7/17/10. \\
-10.1145/3132272.3135074}
+{\emph{ISS '17}}, October 17–20, 2017, Brighton, United Kingdom. \\
+\copyright~2017 Copyright is held by the owner/author(s). \\
+ACM ISBN 978-1-4503-4691-7/17/10. \\
+https://doi.org/10.1145/3132272.3135074}
 %% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
-
-% \CopyrightYear{2017}
-% \setcopyright{rightsretained}
-% \conferenceinfo{ISS '17}{October 17--20, 2017, Brighton, United Kingdom}\isbn{978-1-4503-4691-7/17/10}
-% \doi{https://doi.org/10.1145/3132272.3135074}
 
@@
 \def\plaintitle{Gesture Typing on Virtual Surfaces: Effect of Input Dimensions on Performance}
-\def\plainauthor{Name}
-\def\plainkeywords{Indirect Interaction, Gesture Input, Gesture Keyboard, Mobile, Continuous Interaction, Tabletop, Text Input}
-\def\plaingeneralterms{Documentation, Standardization}
+\def\plainauthor{Antoine Loriette}
+\def\plainkeywords{Indirect Interaction; Gesture Input; Gesture Keyboard; Mobile; Continuous Interaction; Tabletop; Text Input}
+\def\plaingeneralterms{}
 
@@
 
-\category{H.5.m}{Information interfaces and presentation (e.g., HCI)}{Miscellaneous}.
+\category{H.5.2}{Information interfaces and presentation (e.g. HCI)}{User Interfaces}.
 %See \cite{ACMCCS}

### 2017-09-04 paper.tex
remove rule above copyright, typo in biblio, change picture to embed font

@@
 \newcommand{\cdt}[1]{{\small\uppercase{{#1}}}}
-\newcommand{\wpm}{\cdt{wpm} }
+% \newcommand{\smit}[1]{{\textit{{#1}}}}
+% \newcommand{\cdt}[1]{{\uppercase{{#1}}}}
+
+\newcommand{\wpm}{\cdt{wpm}}
 
@@
 % =============================================================================
-The combination of indirect optically tracked input and potentially projected display (“virtual surfaces”) has several interesting properties.
+The combination of indirect optically tracked input and potentially projected display (``virtual surfaces'') has several interesting properties.
 Virtual surfaces, as opposed to touch screens, are well suited to tackle the palm rejection problem and the occlusion problem, offer a choice of size, aspect ratio and texture of the input space, and allow interactions with dirty or wet hands.
@@
 \end{figure}
+
 \begin{figure}
-    \begin{center}
-        \includegraphics[width=\marginparwidth]{figures/system_insitu_0.jpg}
-        \caption{Picture of the user study setup. A tablet device provides the audiovisual feedback while a depth camera mounted on a tripod creates a virtual touch input surface.}
-        \label{fig:system_insitu_0}
-    \end{center}
+    \centering
+    \includegraphics[width=\linewidth]{figures/pointcloud.pdf}
+    \caption{Segmented sensor image with points within the interactive surface (blue), the ROI plane (cyan), the detected fingers (yellow) and the user pointer (red).}~\label{fig:pointcloud}
 \end{figure}
@@
 % =============================================================================
-The envisioned interaction scenario is shown in ~\autoref{fig:banner}.
+The envisioned interaction scenario is shown in~\autoref{fig:banner}.
 A camera overlooks the interaction surface and a computing device performs the hand and fingers tracking as well as the touch detection.
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/roc_auc.pdf}
+    \caption{Performance of the fingertip touch classification with 3-fold cross validation.
+Each split trains on two finger types (among three) and validates against the remaining one.}~\label{fig:roc_auc}
+\end{figure}
+
 The classification features are chosen as 20-bins histograms of z-values of the detected fingertip pointcloud, in red on~\autoref{fig:pointcloud}, which provides some orientation and shape invariance.
@@
 
+\marginpar{
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\linewidth]{figures/pointcloud.pdf}
+%     \caption{Segmented sensor image with points within the interactive surface (blue), the ROI plane (cyan), the detected fingers (yellow) and the user pointer (red).}
+%     \label{fig:pointcloud}
+% \end{figure}
+\begin{figure}
+    \centering
+    % \begin{center}
+    \includegraphics[width=\marginparwidth]{figures/system_insitu_0.jpg}
+    \caption{Picture of the user study setup. A tablet device provides the audiovisual feedback while a depth camera mounted on a tripod creates a virtual touch input surface.}
+    \label{fig:system_insitu_0}
+    % \end{center}
+\end{figure}
+
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\linewidth]{figures/pointcloud.pdf}
+%     \caption{Segmented sensor image with points within the interactive surface (blue), the ROI plane (cyan), the detected fingers (yellow) and the user pointer (red).}~\label{fig:pointcloud}
+% \end{figure}
+}
+
 % =============================================================================
@@
 Accot et al.~\cite{Accot2001} demonstrated a U-shaped curve in performance against input scale.
-Our independent variable were thus the control surface dimensions, which effectively change the control-display gain ($CD_{gain}$) defined as in \cite{Casiez2008} by the the ratio of the pointer velocity to device velocity.
-
-\marginpar{
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{figures/pointcloud.pdf}
-    \caption{Segmented sensor image with points within the interactive surface (blue), the ROI plane (cyan), the detected fingers (yellow) and the user pointer (red).}
-    \label{fig:pointcloud}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{figures/roc_auc.pdf}
-    \caption{Performance of the fingertip touch classification with 3-fold cross validation.
-Each split trains on two finger types (among three) and validates against the remaining one.}~\label{fig:roc_auc}
-\end{figure}
-}
+Our independent variable were thus the control surface dimensions, which effectively change the control-display gain ($CD_{gain}$) defined as in~\cite{Casiez2008} by the the ratio of the pointer velocity to device velocity.
 
@@
 A repeated measures within-subjects design was used.
-There were 3 conditions (\cdt{device}, \cdt{size} and \cdt{orientation}) and 5 level combinations were evaluated, see \autoref{tab:cdt} below for details.
+There were 3 conditions (\cdt{device}, \cdt{size} and \cdt{orientation}) and 5 level combinations were evaluated, see~\autoref{tab:cdt} below for details.
 We adopted a 3-symbols naming convention: the first letter represents the DEVICE, the second marks the ORIENTATION and the number represents the scaling factor as SIZE.
-
-The apparatus, see \autoref{fig:system_insitu_0}, implements the system description above.
-An Intel Realsense camera is used, and the processing is performed on a desktop computer.
-An Android tablet running custom software produces the audiovisual feedback.
-The pointer tracked by the system is defined as the point cloud closest to the camera in the depth direction ($y$ on ~\autoref{fig:pointcloud}); there is no explicit finger or hand modeling performed.
-Instead, the assumption is that the interacting finger is the furthest protruding object from the user.
-
-The task was to write 20 words per level (words taken at random from the most common english words with length between 2 and 5 letters) with maximum 7 attempt to complete 5 correct input for each word.
-The design of the task is similar to \cite{Quinn2016} and allows novice users to focus on the physical execution instead of the shape recollection, effectively emulating a proficient bahavior even for novice users.
-After each level, participants were offered to take a break before moving to the next one.
-Finally, participants were asked for their feedback using the NASA Task Load Index~\cite{Hart1988} (NASA-TLX) to assess the perceived workload of completed level.
-
-The experimental design was thus: 12 participants $\times$ 5 \cdt{level} $\times$ 20 \cdt{word} = 1200 trials.
-For each trial, we had 5 to 7 attemps depending on the error rate, which means a total of 6,000 to 8,400 total samples.
-After the experiment, we actually recorded 6963 samples.
 
@@
 
+The apparatus, see~\autoref{fig:system_insitu_0}, implements the system description above.
+An Intel Realsense camera is used, and the processing is performed on a desktop computer.
+An Android tablet running custom software produces the audiovisual feedback.
+The pointer tracked by the system is defined as the point cloud closest to the camera in the depth direction ($y$ on~\autoref{fig:pointcloud}); there is no explicit finger or hand modeling performed.
+Instead, the assumption is that the interacting finger is the furthest protruding object from the user.
+
+The task was to write 20 words per level (words taken at random from the most common english words with length between 2 and 5 letters) with maximum 7 attempt to complete 5 correct input for each word.
+The design of the task is similar to \cite{Quinn2016} and allows novice users to focus on the physical execution instead of the shape recollection, effectively emulating a proficient bahavior even for novice users.
+After each level, participants were offered to take a break before moving to the next one.
+Finally, participants were asked for their feedback using the NASA Task Load Index~\cite{Hart1988} (NASA-TLX) to assess the perceived workload of completed level.
+
+The experimental design was thus: 12 participants $\times$ 5 \cdt{level} $\times$ 20 \cdt{word} = 1200 trials.
+For each trial, we had 5 to 7 attemps depending on the error rate, which means a total of 6,000 to 8,400 total samples.
+After the experiment, we actually recorded 6963 samples.
+
 % =============================================================================
@@
 % =============================================================================
-The dependent variables we analyse are the success rate, the time taken per trial and the trace data when available.
-From these, we computed the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in~\cite{Markussen2014}, according to the following formula: $WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string and $s$ is time in seconds.
-
-We found one input word (\cdt{lay}) presented some very unusual behavior.
-Its error rate was at $89.5\%$ while the \cdt{WORD} mean was $20.1\%$ and no other word had an error rate higher than $30\%$.
-The reason was that the recogniser promoted words of higher prior probability in the language model, \textit{``Larry''}, \textit{``last''} or \textit{``Katy''}.
-For the all subsequent analysis of error rate, \cdt{lay} was removed from the dataset.
-
-\subsection{Effect on error rate}
-A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77, p < 0.0001$) on error rate with mean value for \cdt{TP1} and \cdt{OP1} equal to 6.2\% and 26.1\%, respectively, see \autoref{fig:err_DEVICE_SHAPE}.
-A statistical analysis also showed a significant main effect of \cdt{SIZE} ($F_{2,22} = 10.99, p < 0.001$) on the error rate.
-Finally, an ANOVA on \cdt{TP1}, \cdt{OP2}, \cdt{OP4} and \cdt{OL2} did not show a significant main effect ($p = 0.11$) even though \cdt{OPTICAL} is on average higher than \cdt{TABLET}.
-In other words, among all levels of the experiment, only \cdt{OP1} shows a significant higher error rate.
-
 \marginpar{
@@
 
+The dependent variables we analyse are the success rate, the time taken per trial and the trace data when available.
+From these, we computed the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in~\cite{Markussen2014}, according to the following formula: $WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string and $s$ is time in seconds.
+
+We found one input word (\cdt{lay}) presented some very unusual behavior.
+Its error rate was at $89.5\%$ while the \cdt{WORD} mean was $20.1\%$ and no other word had an error rate higher than $30\%$.
+The reason was that the recogniser promoted words of higher prior probability in the language model, \textit{``Larry''}, \textit{``last''} or \textit{``Katy''}.
+For the all subsequent analysis of error rate, \cdt{lay} was removed from the dataset.
+
+\subsection{Effect on error rate}
+A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77$, $p < 0.0001$) on error rate with mean value for \cdt{TP1} and \cdt{OP1} equal to 6.2\% and 26.1\%, respectively, see~\autoref{fig:err_DEVICE_SHAPE}.
+A statistical analysis also showed a significant main effect of \cdt{SIZE} ($F_{2,22} = 10.99$, $p < 0.001$) on the error rate.
+Finally, an ANOVA on \cdt{TP1}, \cdt{OP2}, \cdt{OP4} and \cdt{OL2} did not show a significant main effect ($p = 0.11$) even though \cdt{OPTICAL} is on average higher than \cdt{TABLET}.
+In other words, among all levels of the experiment, only \cdt{OP1} shows a significant higher error rate.
+
 \subsection{Effect on input rate}
-A statistical analysis showed a significant main effect of \cdt{DEVICE} ($F_{1,11} = 90.15, p < 0.0001$) on input rate, see \autoref{fig:wpm_DEVICE_SHAPE}, with mean values for \cdt{TP1} and \cdt{OP1} equal to $29.2 WPM$ and $13.7 WPM$, respectively.
-We also looked at pairwise comparison\footnote{Post-hoc analysis were adjusted using Bonferroni correction.} for \cdt{OPTICAL} and could not find a statistical difference in the mean input rate.
+A statistical analysis showed a significant main effect of \cdt{DEVICE} ($F_{1,11} = 90.15$, $p < 0.0001$) on input rate, see~\autoref{fig:wpm_DEVICE_SHAPE}, with mean values for \cdt{TP1} and \cdt{OP1} equal to $29.2$ WPM and $13.7$ WPM, respectively.
+We also looked at pairwise comparison
+% \footnote{Post-hoc analysis were adjusted using Bonferroni correction.}
+for \cdt{OPTICAL} and could not find a statistical difference in the mean input rate.
 The input rate achieved by the participants in \cdt{TP1} is in-line with what can be expected from novice users after the time of the experiments~\cite{Kristensson2004}.
@@
 The main result is the average pointer speed in display space at $303.75 pixel/s$, higher than the PORTRAIT orientation at an average $228 pixel/s$.
-It is important to keep in mind that the keyboard area in LANDSCAPE ($61cm^2$), bigger than PORTRAIT ($44cm^2$), which can explain the faster pointer speed but is not responsible for a higher input rate due to longer traces to produces.
+It is important to keep in mind that the keyboard area in LANDSCAPE ($61 cm^2$), bigger than PORTRAIT ($44 cm^2$), which can explain the faster pointer speed but is not responsible for a higher input rate due to longer traces to produces.
 
@@
 Accot et al.~\cite{Accot2001} showed that task with high index of difficulty would exhibits a u-shaped curved with size.
-Since we did not observe such a behavior, this puts in question whether gesture typing for novice users is “hard enough”.
+Since we did not observe such a behavior, this puts in question whether gesture typing for novice users is ``hard enough''.
 
